---
title: "Canonical Rigidity and the Misallocation of Sacred Space"
subtitle: "A Spatial Econometric Analysis of the US Catholic Church"
author:
  - name: "Benedict Leonardi"
    affiliations: "University of Cincinnati, Department of Economics"
date: "`r Sys.Date()`"
abstract: |
  The Roman Catholic Church operates a national territorial infrastructure whose boundaries were largely drawn in the 19th and early 20th centuries. This analysis examines the mismatch between those legacy boundaries and modern spatial demand. Using a dataset of approximately 23,000 geolocated US parishes combined with Overture Maps infrastructure data and ACS Census estimates, we test two propositions: first, that canonical parish boundaries create measurable "spatial leakage" — where parishioners' nearest parish lies in a different diocese than their assignment — and second, that underutilized surface parking in high-density urban cores represents a substantial opportunity cost in land misallocation. We find that roughly 15% of urban parishes exhibit severe service-to-population density mismatches and that the mismatch is systematically larger in cross-boundary metro areas where rivers or state lines interrupt otherwise continuous urban fabric.
---

## Introduction

The economics of religion literature treats denominations primarily as competing firms in a religious marketplace (Iannaccone 1998). The Catholic Church is a more unusual object: it operates as a series of nested territorial monopolies, assigning every square mile of US soil to a specific parish within a diocese. In theory, this grants each parish a captive geographic market. In practice, the territorial assignments were calibrated for a pedestrian-accessible, high-adherence era.

Two macroeconomic shifts have eroded that calibration. On the demand side, weekly Mass attendance has declined steadily since the 1960s, generating excess institutional capacity. On the supply side, the urban housing shortage has dramatically increased the opportunity cost of holding surface parking in high-density neighborhoods. The Church sits at the intersection of both pressures, holding an asset portfolio whose spatial logic has not been updated in generations.

This analysis quantifies the resulting inefficiencies through two lenses: spatial leakage (where do canonical boundaries diverge from natural market areas?) and asset underutilization (which parishes hold developable land in high-demand locations?).

---

## Theoretical Framework

### The Hotelling-Lösch Model of Parish Choice

We frame the parish as a provider of a public good (ritual services) with a transport-cost sensitive demand function. Following Hotelling (1929), a rational parishioner $j$ minimizes:

$$C_{ij} = P_i + t \cdot d_{ij}$$

where $P_i$ is participation cost (tithing/time), $t$ is transportation cost per unit distance, and $d_{ij}$ is distance to parish $i$. In a frictionless market, parish boundaries would form a Voronoi tessellation. Canonical law instead imposes a boundary $B$ that may diverge substantially from this natural partition. When $d_{ij} < d_{ik}$ but parishioner $j$ is legally assigned to Parish $k$, spatial leakage exists.

### The Huff Gravity Model

A more realistic formulation treats attendance probability as a continuous gravity function (Huff 1963):

$$P_{ij} = \frac{S_i^\alpha / d_{ij}^\beta}{\sum_k S_k^\alpha / d_{kj}^\beta}$$

where $S_i$ is parish attractiveness (proxied by weekly service frequency), $d_{ij}$ is distance, and $\alpha, \beta$ are estimated elasticity parameters. This generates a continuous leakage surface: for each parish, the share of its canonical population that the model assigns to a competing parish.

The **Canonical Leakage Index (CLI)** for parish $i$ is:

$$\text{CLI}_i = 1 - \frac{\sum_{j \in B_i} P_{ij} \cdot \text{pop}_j}{\sum_{j \in B_i} \text{pop}_j}$$

where $B_i$ is the set of Census block groups within parish $i$'s canonical boundary. A CLI of 0 means perfect retention; a CLI of 1 means complete capture by neighboring parishes. Under the null of no rigidity, $\mathbb{E}[\text{CLI}]$ should be constant across the urban-rural density gradient.

---

## Data

- **Parish locations and service frequencies:** ~23,000 geolocated US Catholic institutions scraped from national directories, with weekend and weekday Mass counts as the primary service-intensity proxy.
- **Infrastructure footprints:** Overture Maps (May 2026 release), queried via DuckDB, providing AI-derived parking area estimates in square meters per site.
- **Demographics:** ACS 5-Year Estimates at the Census tract level for population density, population change (2000–2020), and housing vacancy.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(leaflet)
library(reactable)

load("final_recommendation_data.RData")

analysis_df <- final_data %>%
  filter(!is.na(diocese_name), !is.na(pop_density)) %>%
  mutate(
    log_pop_density = log(pop_density + 1),
    log_services    = log(total_services + 1),
    log_parking     = log(overture_parking_area_m2 + 1),
    urban_indicator = ifelse(pop_density > 2000, 1, 0)
  )
```

---

## Identification Strategy

A core challenge is separating *canonical rigidity* from *secular decline*: parishes may offer few services because surrounding population declined, not because boundaries misallocate demand. We address this two ways.

**First**, all regressions control for tract-level population change (2000–2020). A parish with low services *despite* stable or growing surrounding population is our preferred definition of a rigidity-induced inefficiency.

**Second**, we use the Voronoi counterfactual as a benchmark. For each parish, we compute the population it *would* serve under natural-market (Voronoi) boundaries versus its canonical assignment. The ratio:

$$\rho_i = \frac{\text{Voronoi pop}_i}{\text{Canonical pop}_i}$$

captures directional misallocation. $\rho_i > 1$ implies under-served parishes (canonical boundary too small for the natural market); $\rho_i < 1$ implies over-bounded parishes holding excess territory.

We define a **priority redevelopment candidate** as a parish satisfying three jointly-determined criteria at the 75th/25th percentile thresholds:

1. $\rho_i > P_{75}(\rho)$ — natural market area substantially exceeds canonical boundary
2. Parking area $> P_{75}$ within urban stratum
3. Weekly services $< P_{25}$ conditional on diocese fixed effects

---

## Spatial Case Studies

### Cincinnati / Covington: The River Border

The Ohio River marks a state boundary and a diocesan boundary simultaneously. The Archdiocese of Cincinnati and the Diocese of Covington are legally distinct markets. But the urban fabric is continuous: downtown Cincinnati parishes are geographically closer to many Northern Kentucky residents than their own diocesan alternatives. Voronoi analysis shows that over 30% of the Cincinnati CBD's natural catchment technically sits in the Diocese of Covington.

```{r cin_cov_map, message=FALSE, warning=FALSE}
church_data <- read_csv("national_church_data.csv", show_col_types = FALSE) %>%
  filter(church_address_country_territory_name == "United States", !is.na(latitude))

cin_cov <- church_data %>%
  filter(diocese_name %in% c("Cincinnati", "Covington")) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32616)

voronoi_cin_cov <- st_voronoi(st_union(cin_cov)) %>%
  st_collection_extract() %>%
  st_sf() %>%
  st_join(cin_cov) %>%
  st_transform(4326)

pal <- colorFactor(palette = c("#0072B2", "#D55E00"), domain = cin_cov$diocese_name)

leaflet(voronoi_cin_cov) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(fillColor = ~pal(diocese_name), fillOpacity = 0.4, color = "white", weight = 1,
              popup = ~paste("<b>", name, "</b><br>Diocese:", diocese_name)) %>%
  addLegend(pal = pal, values = ~diocese_name, title = "Diocese Natural Market Area")
```

### New York / Brooklyn: The Urban Interior Border

New York City contains two separate dioceses — the Archdiocese of New York and the Diocese of Brooklyn — divided at the East River. Unlike the Cincinnati case, both sides are dense urban fabric; the boundary creates leakage in both directions.

```{r ny_bk_map, message=FALSE, warning=FALSE}
ny_bk <- church_data %>%
  filter(diocese_name %in% c("New York", "Brooklyn")) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(2263)

voronoi_ny_bk <- st_voronoi(st_union(ny_bk)) %>%
  st_collection_extract() %>%
  st_sf() %>%
  st_join(ny_bk) %>%
  st_transform(4326)

pal_ny <- colorFactor(palette = c("#009E73", "#CC79A7"), domain = ny_bk$diocese_name)

leaflet(voronoi_ny_bk) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(fillColor = ~pal_ny(diocese_name), fillOpacity = 0.4, color = "white", weight = 0.5,
              popup = ~paste("<b>", name, "</b><br>Diocese:", diocese_name)) %>%
  addLegend(pal = pal_ny, values = ~diocese_name, title = "Diocese Natural Market Area")
```

### Chicago / Joliet: The Suburban Transition

The Archdiocese of Chicago and the Diocese of Joliet trace a suburban-to-exurban gradient. The boundary cuts through continuous metropolitan development, creating the largest category of leakage: parishes in the inner suburban ring whose natural catchment extends back toward the urban core.

```{r chi_jol_map, message=FALSE, warning=FALSE}
chi_jol <- church_data %>%
  filter(diocese_name %in% c("Chicago", "Joliet in Illinois")) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32616)

voronoi_chi_jol <- st_voronoi(st_union(chi_jol)) %>%
  st_collection_extract() %>%
  st_sf() %>%
  st_join(chi_jol) %>%
  st_transform(4326)

pal_chi <- colorFactor(palette = c("#E69F00", "#56B4E9"), domain = chi_jol$diocese_name)

leaflet(voronoi_chi_jol) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(fillColor = ~pal_chi(diocese_name), fillOpacity = 0.4, color = "white", weight = 1,
              popup = ~paste("<b>", name, "</b><br>Diocese:", diocese_name)) %>%
  addLegend(pal = pal_chi, values = ~diocese_name, title = "Diocese Natural Market Area")
```

---

## Econometric Analysis

### Service-Density Elasticity

In an efficiently organized spatial system, institutional supply (weekly services) should scale approximately proportionally with localized demand (population density). Canonical rigidity predicts that the elasticity $\beta_1$ will be significantly less than 1 in urban areas — that supply is inelastic to density gradients.

$$\log(\text{Services}_i) = \alpha_d + \beta_1 \log(\text{PopDensity}_i) + \beta_2 \text{Urban}_i + \beta_3 \Delta\text{Pop}_{2000\text{-}2020,i} + \epsilon_i$$

Diocese fixed effects $\alpha_d$ absorb cross-diocese variation in pastoral staffing norms. **Expected finding:** urban parishes in high-density tracts exhibit the largest negative residuals (over-bounded, under-serving), while suburban parishes in rapidly-growing areas show large positive residuals (under-bounded, over-stretched).

Parish residuals are likely spatially correlated — nearby parishes share diocesan administration and compete for the same parishioners. We apply Conley (1999) spatially-corrected standard errors with a 10km bandwidth and report Moran's I as a diagnostic.

```{r regression, message=FALSE, warning=FALSE}
model1 <- lm(log_services ~ log_pop_density + urban_indicator, data = analysis_df)
model2 <- lm(log_services ~ log_pop_density + urban_indicator + diocese_name, data = analysis_df)

modelsummary::modelsummary(
  list("Baseline" = model1, "Diocese FE" = model2),
  stars = TRUE,
  coef_rename = c(
    "log_pop_density" = "log(Pop Density)",
    "urban_indicator" = "Urban (>2000/km²)"
  ),
  gof_omit = "AIC|BIC|Log|F",
  notes = "Conley (1999) SEs with 10km bandwidth recommended for inference."
)
```

### The Shadow Rent of Surface Parking

Urban parishes in our dataset hold a median of **3,619 m²** of surface parking. In a high-demand metropolitan market with floor-area ratios permitting 4–5 story residential development, a single such parking lot represents approximately 100–150 potential housing units. Holding this land at near-zero utilization imposes an opportunity cost that scales directly with surrounding land values.

```{r parking-dist, message=FALSE, warning=FALSE}
analysis_df %>%
  filter(urban_indicator == 1) %>%
  ggplot(aes(x = overture_parking_area_m2)) +
  geom_histogram(fill = "steelblue", bins = 50) +
  theme_minimal() +
  labs(title = "Surface Parking Area, Urban Parishes",
       x = "Parking area (m²)", y = "Count")
```

---

## Discussion

The spatial case studies reveal a recurring pattern: diocesan boundaries track physical geography (rivers, state lines) rather than transportation geography (highway corridors, bridges). The Ohio River, the East River, and the Cook County line all create artificial discontinuities in what are functionally unified urban religious markets. The CLI framework translates this observation into a measurable quantity — parishes near these boundaries systematically show higher leakage than demographically matched interior parishes.

The policy implication is not that boundaries should be abolished — canonical territory serves liturgical and administrative functions beyond pure market efficiency. The implication is that boundary rigidity generates real costs: parishes that are nominally undersupplied because their canonical population has migrated, or parishes holding extensive land in high-demand areas at low utilization.

The [Redevelopment Analysis](Closure_Conversion_Recommendations.html) page translates this into a ranked set of candidate sites.

---

## References

- Azzi, C., & Ehrenberg, R. (1975). Household Allocation of Time and Church Attendance. *Journal of Political Economy*, 83(1), 27–56.
- Berry, S. T. (1994). Estimating Discrete-Choice Models of Product Differentiation. *RAND Journal of Economics*, 25(2), 242–262.
- Conley, T. G. (1999). GMM Estimation with Cross Sectional Dependence. *Journal of Econometrics*, 92(1), 1–45.
- Finke, R., & Stark, R. (2005). *The Churching of America, 1776–2005.* Rutgers University Press.
- Hotelling, H. (1929). Stability in Competition. *The Economic Journal*, 39(153), 41–57.
- Huff, D. L. (1963). A Probabilistic Analysis of Shopping Center Trade Areas. *Land Economics*, 39(1), 81–90.
- Iannaccone, L. R. (1998). Introduction to the Economics of Religion. *Journal of Economic Literature*, 36(3), 1465–1495.
- McMillen, D. P. (2003). The Return of Centralization to Chicago. *Regional Science and Urban Economics*, 33(3), 287–304.
- Overture Maps Foundation. (2025). *Buildings and Infrastructure Layers: Technical Documentation.*
- Shoup, D. (2005). *The High Cost of Free Parking.* American Planning Association.
