---
title: "Canonical Rigidity and the Misallocation of Sacred Space"
subtitle: "Girard, Lösch, and the Territorial Logic of the American Catholic Church"
author:
  - name: "Benedict Leonardi"
    affiliations: "University of Cincinnati, Department of Economics"
date: "`r Sys.Date()`"
bibliography: references.bib
abstract: |
  The Roman Catholic Church operates a national territorial infrastructure — approximately 23,000 parishes — whose boundaries were drawn for a pedestrian, high-adherence 19th century. We argue that this infrastructure exhibits two interlocking pathologies. The first is economic: canonical boundaries diverge from the Löschian natural market partition, producing measurable spatial leakage (the Canonical Leakage Index) and chronic asset underutilization. The second is anthropological: following Girard, the diocesan institution responds to territorial crisis not by reorganizing the map but by sacrificing individual parishes — designating *pharmakoi* whose closure restores apparent order without touching the underlying misallocation. We trace both pathologies through a national dataset of 23,000 geolocated parishes, two deep-dive dioceses (Detroit and Cincinnati), and 26 years of sacramental panel data. Detroit, in full crisis, illustrates the complete Girardian sequence: mimetic undifferentiation → sacrificial selection → spatial void. Cincinnati, earlier in the same arc, provides the pre-crisis control, validating the structural pattern while isolating the AOD amplifier specific to Detroit. Together they establish that the institution preserves its canonical map by consuming its weakest nodes — and that the cost of that preservation is borne spatially, by the communities that surround the sacrifice.
---

## I. Introduction

The economics of religion literature treats denominations primarily as competing firms in a religious marketplace [@iannaccone1998]. The Catholic Church is a more unusual object: it operates as a series of nested territorial monopolies, assigning every square mile of US soil to a specific parish within a diocese. In theory, this grants each parish a captive geographic market. In practice, the territorial assignments were calibrated for a pedestrian-accessible, high-adherence era, and two macroeconomic shifts have eroded that calibration. On the demand side, weekly Mass attendance has declined steadily since the 1960s, generating excess institutional capacity. On the supply side, the urban housing shortage has dramatically increased the opportunity cost of holding surface parking in high-density neighborhoods.

The economic framing captures what the Church is trying to do — allocate ritual services across space — but not why it consistently fails to adapt the map through which it does so. That question requires an anthropological supplement. We argue, following @girard1972, that the persistence of canonical rigidity is not a coordination failure. It is a sacrificial logic: when the territorial order that constitutes the diocese as an institution is threatened by demographic change, the institution responds not by reorganizing the order but by consuming its weakest nodes. The parish that is closed is the *pharmakos* — the scapegoat whose expulsion temporarily restores coherence to a system whose underlying misallocation is left intact.

This paper layers Girard onto Lösch. The Löschian framework [@losch1954] tells us what efficient allocation of sacred services across space would look like and measures the gap between that ideal and the canonical reality. The Girardian framework tells us why the institution will not close that gap through internal reform alone, and what the sacrifice costs the communities that absorb the void. The paper proceeds through four movements: theoretical (§II), historical (§III), cross-sectional (§§IV–V), and longitudinal (§§VI–VIII), culminating in a land-economics assessment of what the sacrifice leaves behind (§IX).

---

## II. Theoretical Framework

### IIa. The Löschian Parish: Hotelling, Huff, and the Canonical Leakage Index

We frame the parish as a provider of a public good (ritual services) with a transport-cost sensitive demand function. Following @hotelling1929, a rational parishioner $j$ minimizes:

$$C_{ij} = P_i + t \cdot d_{ij}$$

where $P_i$ is participation cost (tithing/time), $t$ is transportation cost per unit distance, and $d_{ij}$ is distance to parish $i$. In a frictionless market, parish boundaries would form a Voronoi tessellation — each parishioner assigned to their nearest parish. Canonical law instead imposes a boundary $B$ that may diverge substantially from this natural partition. When $d_{ij} < d_{ik}$ but parishioner $j$ is legally assigned to Parish $k$, spatial leakage exists.

A more realistic formulation treats attendance probability as a continuous gravity function [@huff1963]:

$$P_{ij} = \frac{S_i^\alpha / d_{ij}^\beta}{\sum_k S_k^\alpha / d_{kj}^\beta}$$

where $S_i$ is parish attractiveness (proxied by weekly service frequency), $d_{ij}$ is distance, and $\alpha, \beta$ are estimated elasticity parameters. This generates a continuous leakage surface.

The **Canonical Leakage Index (CLI)** for parish $i$ is:

$$\text{CLI}_i = 1 - \frac{\sum_{j \in B_i} P_{ij} \cdot \text{pop}_j}{\sum_{j \in B_i} \text{pop}_j}$$

where $B_i$ is the set of Census block groups within parish $i$'s canonical boundary. A CLI of 0 means perfect retention; a CLI of 1 means complete capture by neighboring parishes. Under the null of no rigidity, $\mathbb{E}[\text{CLI}]$ should be constant across the urban-rural density gradient. The observed pattern — CLI rising sharply with urban density — is the cross-sectional signature of canonical rigidity.

### IIb. The Girardian Diocese: Mimetic Crisis and the Pharmakos

@girard1972 argues that social order is constituted through differentiation — the drawing of distinctions that prevent the undifferentiated mimetic rivalry he calls the "crisis of differences." Sacred boundaries are the primary instrument of that differentiation. In the Catholic diocesan context, canonical territorial boundaries are not merely administrative lines: they are the spatial expression of a differentiating order that assigns every Catholic body to a specific sacred jurisdiction, distinguishing each parish from its neighbors through the particularity of territory, history, and community.

This differentiating order was stable so long as the parish's canonical boundary and its social community were roughly coextensive — which they were in the pedestrian city of the early 20th century. The parish was the neighborhood; the neighborhood was the parish; the boundary between them was both canonical and social. Postwar automobile mobility dissolved this coextension without dissolving the canonical map. Populations migrated outward; boundaries did not. The canonical distinctions that maintained order now float free of the social reality they were meant to organize. This is what Girard would recognize as mimetic undifferentiation: the boundaries are still drawn, but the differences they mark have ceased to correspond to anything outside themselves.

The institution's characteristic response to a crisis of differences is the scapegoat mechanism [@girard1986]. Rather than reorganize the differentiating order — which would require acknowledging that the order itself is contingent — the institution designates a *pharmakos* whose expulsion restores apparent coherence. The scapegoat is selected not because it is the cause of the crisis, but because it is sufficiently marginal and sufficiently distinguishable to serve as its symbolic container. In the diocesan restructuring process, the pharmakos is the parish marked for closure. It is selected on attendance and financial metrics — measures of its own weakness — while the canonical map that produced its weakness is left untouched. Each closure is announced as a rational response to demographic reality. What it actually accomplishes is the preservation of the territorial fiction by consuming one of its weakest nodes.

The empirical implication is testable: if the diocese operates sacrificial logic, then high-CLI parishes — those whose canonical boundaries have become most fictive, most divorced from their actual social catchment — should face elevated closure risk, controlling for their absolute scale of decline. The misallocation itself should predict the sacrifice.

---

## III. Historical Origins: The Making of the Misallocation

The CLI framework measures the *current* state of canonical rigidity but does not explain why the misallocation is deepest in the metros where it is most entrenched. That explanation requires a historical lens. Suburbanization is the triggering event of the mimetic crisis: the moment when the pedestrian map of the early 20th century stopped matching the automobile geography that replaced it. The congregation-trend data show that high-CLI metros are precisely those with the deepest pre-suburban Catholic institutional roots — cities where the parish network was so dense before the postwar wave that the institution never fully relocated outward with its membership. The Q5 metros (Boston, Chicago, Detroit, Milwaukee, Providence) had the most differentiated pre-suburban parish geography to lose, which is why the mimetic crisis is deepest there.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(leaflet)
library(reactable)
library(readxl)
library(scales)
library(fixest)
library(spdep)
library(broom)
library(tidycensus)

okabe_ito <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
               "#0072B2", "#D55E00", "#CC79A7", "#000000")

load("final_recommendation_data.RData")

analysis_df <- final_data %>%
  filter(!is.na(diocese_name), !is.na(pop_density)) %>%
  mutate(
    log_pop_density = log(pop_density + 1),
    log_services    = log(total_services + 1),
    log_parking     = log(overture_parking_area_m2 + 1),
    urban_indicator = ifelse(pop_density > 2000, 1, 0)
  )

church_data <- read_csv("national_church_data.csv", show_col_types = FALSE) %>%
  filter(
    church_address_country_territory_name == "United States",
    !is.na(latitude), !is.na(longitude),
    is.finite(latitude), is.finite(longitude),
    latitude > 18, latitude < 72, longitude > -180, longitude < -60,
    church_type_name %in% c("Parish", "Cathedral", "Basilica", "Mission", "Shrine")
  )

worship <- read_csv("national_worship_times.csv", show_col_types = FALSE) %>%
  group_by(psuedo_id, service_typename) %>%
  summarize(ct = n(), .groups = "drop") %>%
  pivot_wider(names_from = service_typename, values_from = ct, values_fill = 0L) %>%
  rename_with(~ paste0("n_", str_to_lower(str_replace_all(., " ", "_"))),
              .cols = -psuedo_id)
church_data <- church_data %>%
  left_join(worship, by = "psuedo_id") %>%
  mutate(n_weekend     = coalesce(n_weekend, 0L),
         n_confessions = coalesce(n_confessions, 0L))

nchs <- read_excel("NCHSURCodes2013.xlsx") %>%
  transmute(
    county_fips = str_pad(as.character(`FIPS code`), 5, pad = "0"),
    cbsa_title  = `CBSA title`,
    ur_code     = `2013 code`,
    ur_label    = case_when(
      ur_code == 1 ~ "Large central metro",
      ur_code == 2 ~ "Large fringe metro",
      ur_code == 3 ~ "Medium metro",
      ur_code == 4 ~ "Small metro",
      ur_code == 5 ~ "Micropolitan",
      ur_code == 6 ~ "Noncore"
    ),
    ur_type = case_when(
      ur_code == 1 ~ "Urban core",
      ur_code == 2 ~ "Suburban",
      ur_code %in% 3:4 ~ "Smaller metro",
      ur_code %in% 5:6 ~ "Rural"
    )
  )

zip_county <- read_csv("zip_county_crosswalk.csv", show_col_types = FALSE) %>%
  transmute(
    zip         = str_pad(as.character(ZCTA5), 5, pad = "0"),
    county_fips = str_pad(as.character(GEOID), 5, pad = "0"),
    area_pct    = ZPOPPCT
  ) %>%
  group_by(zip) %>%
  slice_max(area_pct, n = 1, with_ties = FALSE) %>%
  ungroup()

county_pop <- get_acs("county", "B01003_001", year = 2019) %>%
  transmute(county_fips = GEOID, county_pop = estimate)

church_data <- church_data %>%
  mutate(zip = str_pad(str_sub(church_address_postal_code, 1, 5), 5, pad = "0")) %>%
  left_join(zip_county,  by = "zip") %>%
  left_join(county_pop,  by = "county_fips") %>%
  left_join(nchs,        by = "county_fips")

arda_raw <- read_excel("arda_longitudinal_1980_2010.xlsx") %>%
  mutate(county_fips = str_pad(as.character(FIPSMERG), 5, pad = "0"))

cath_long <- arda_raw %>%
  filter(GRPCODE == "081") %>%
  transmute(county_fips, year = YEAR, totpop = TOTPOP,
            cath_cong = coalesce(CONGREG, 0L),
            cath_adh  = coalesce(ADHERENT, 0L),
            pct_cath  = cath_adh / totpop)

all_cong <- arda_raw %>%
  group_by(county_fips, year = YEAR) %>%
  summarize(tot_cong = sum(coalesce(CONGREG, 0L), na.rm = TRUE),
            totpop   = first(TOTPOP), .groups = "drop")

county_cath <- cath_long %>%
  pivot_wider(id_cols = county_fips, names_from = year,
              values_from = c(totpop, cath_adh, pct_cath, cath_cong),
              names_glue = "{.value}_{year}") %>%
  mutate(cath_share_change_80_10 = pct_cath_2010 - pct_cath_1980,
         cath_quintile_1980      = ntile(pct_cath_1980, 5))

county_data <- cath_long %>%
  left_join(all_cong %>% select(county_fips, year, tot_cong), by = c("county_fips","year")) %>%
  left_join(nchs, by = "county_fips") %>%
  filter(!is.na(ur_type))

church_data <- church_data %>%
  left_join(county_cath, by = "county_fips")

metro_cath_1980 <- county_data %>%
  filter(year == 1980, !is.na(cbsa_title)) %>%
  group_by(cbsa_title) %>%
  summarize(metro_pct_cath_1980 = weighted.mean(pct_cath, totpop, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(metro_cath_q = ntile(metro_pct_cath_1980, 5))

church_data <- church_data %>%
  left_join(metro_cath_1980, by = "cbsa_title")
```

### Catholic Congregation Trends by Urban-Rural Type, 1980–2010

The ARDA Longitudinal Religious Congregations and Membership file (1980–2010) provides Catholic congregation counts at the county level across four decades. Joined to the NCHS 2013 urban-rural classification, it allows direct measurement of how the Catholic institutional footprint shifted across geographies relative to population.

```{r cong-trend, message=FALSE, warning=FALSE}
cong_ur <- county_data %>%
  group_by(ur_type, year) %>%
  summarize(
    cath_cong = sum(cath_cong, na.rm = TRUE),
    tot_cong  = sum(tot_cong,  na.rm = TRUE),
    totpop    = sum(totpop,    na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  group_by(ur_type) %>%
  mutate(
    idx_cong    = cath_cong / cath_cong[year == 1980] * 100,
    idx_pop     = totpop    / totpop[year == 1980]    * 100,
    cong_vs_pop = idx_cong / idx_pop * 100
  ) %>%
  ungroup()
```

```{r cong-index-plot, message=FALSE, warning=FALSE}
cong_ur %>%
  select(ur_type, year, idx_cong, idx_pop) %>%
  pivot_longer(c(idx_cong, idx_pop), names_to = "series", values_to = "index") %>%
  mutate(
    ur_type = fct_relevel(ur_type, "Urban core", "Suburban", "Smaller metro", "Rural"),
    series  = if_else(series == "idx_cong", "Catholic congregations", "Total population")
  ) %>%
  ggplot(aes(x = year, y = index, color = series, linetype = series)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dotted", color = "grey60") +
  facet_wrap(~ ur_type, nrow = 1) +
  scale_color_manual(values = c("Catholic congregations" = "#c0392b",
                                "Total population"       = "#2980b9")) +
  scale_linetype_manual(values = c("Catholic congregations" = "solid",
                                   "Total population"       = "dashed")) +
  labs(
    title    = "Catholic congregations did not follow the suburban population",
    subtitle = "1980 = 100; ARDA congregation counts vs. Census population by NCHS urban-rural type",
    x = NULL, y = "Index (1980 = 100)", color = NULL, linetype = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```

### Anchoring to 1952: The Pre-Suburbanization Baseline

```{r load-1952, message=FALSE, warning=FALSE}
ccm_1952 <- read_excel("arda_ccm_1952_county.xls") %>%
  transmute(
    county_fips   = str_pad(as.character(STCODE * 1000 + CCODE), 5, pad = "0"),
    cath_memb_52  = coalesce(CATH_M, 0),
    cath_cong_52  = coalesce(CATH_C, 0),
    totpop_52     = TOTPOP,
    pct_cath_1952 = cath_memb_52 / totpop_52
  )

county_cath <- county_cath %>% left_join(ccm_1952, by = "county_fips")

cat("1952 coverage:", sum(!is.na(county_cath$pct_cath_1952)), "counties\n")
cat("Correlation pct_cath_1952 ~ pct_cath_1980:",
    round(cor(county_cath$pct_cath_1952, county_cath$pct_cath_1980,
              use = "complete.obs"), 3), "\n")
```

County-level Catholic concentration is highly stable between 1952 and 1980, validating the earlier year as a true pre-suburbanization baseline.

```{r full-timeseries, message=FALSE, warning=FALSE}
cong_1952 <- ccm_1952 %>%
  left_join(nchs %>% select(county_fips, ur_type), by = "county_fips") %>%
  filter(!is.na(ur_type)) %>%
  group_by(ur_type) %>%
  summarize(cath_cong = sum(cath_cong_52, na.rm = TRUE),
            totpop    = sum(totpop_52,    na.rm = TRUE),
            .groups   = "drop") %>%
  mutate(year = 1952)

cong_full <- bind_rows(cong_1952, cong_ur %>% select(ur_type, year, cath_cong, totpop)) %>%
  group_by(ur_type) %>%
  mutate(
    idx_cong = cath_cong / cath_cong[year == 1952] * 100,
    idx_pop  = totpop    / totpop[year == 1952]    * 100,
    cong_vs_pop = idx_cong / idx_pop * 100
  ) %>%
  ungroup()

cong_full %>%
  mutate(ur_type = fct_relevel(ur_type, "Urban core", "Suburban", "Smaller metro", "Rural")) %>%
  ggplot(aes(x = year, y = cong_vs_pop, color = ur_type, group = ur_type)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = okabe_ito) +
  scale_x_continuous(breaks = c(1952, 1980, 1990, 2000, 2010)) +
  annotate("rect", xmin = 1945, xmax = 1975, ymin = -Inf, ymax = Inf,
           alpha = 0.06, fill = "steelblue") +
  annotate("text", x = 1952, y = 67, label = "Peak\nsuburbanization\nwave",
           hjust = 0, size = 2.8, color = "steelblue4") +
  labs(
    title    = "Catholic institutional lag relative to population, 1952–2010",
    subtitle = "Congregation / population index (1952 = 100); shaded = peak suburbanization era",
    x = NULL, y = "Congregation / population index", color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### The Q5 Metros: Where the Crisis Is Deepest

```{r catholic-cities-q, message=FALSE, warning=FALSE}
metro_cath_1952 <- ccm_1952 %>%
  left_join(nchs %>% select(county_fips, cbsa_title), by = "county_fips") %>%
  filter(!is.na(cbsa_title)) %>%
  group_by(cbsa_title) %>%
  summarize(
    metro_pct_cath_1952  = weighted.mean(pct_cath_1952, totpop_52, na.rm = TRUE),
    metro_cath_cong_1952 = sum(cath_cong_52, na.rm = TRUE),
    metro_pop_1952       = sum(totpop_52, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(metro_pop_1952 >= 250000) %>%
  mutate(cath_q_1952 = ntile(metro_pct_cath_1952, 5))

county_q52 <- county_data %>%
  filter(!is.na(cbsa_title), ur_type %in% c("Urban core", "Suburban")) %>%
  left_join(metro_cath_1952, by = "cbsa_title") %>%
  filter(!is.na(cath_q_1952))

shift_q52 <- county_q52 %>%
  group_by(cath_q_1952, ur_type, year) %>%
  summarize(cath_cong = sum(cath_cong, na.rm = TRUE),
            totpop    = sum(totpop,    na.rm = TRUE),
            .groups   = "drop") %>%
  group_by(cath_q_1952, ur_type) %>%
  mutate(idx_cong    = cath_cong / cath_cong[year == 1980] * 100,
         idx_pop     = totpop    / totpop[year == 1980]    * 100,
         cong_vs_pop = idx_cong / idx_pop * 100) %>%
  ungroup()

shift_q52 %>%
  mutate(
    ur_type     = fct_relevel(ur_type, "Urban core", "Suburban"),
    quintile_lab = paste0("Q", cath_q_1952,
                          if_else(cath_q_1952 == 1, " (least Catholic 1952)",
                          if_else(cath_q_1952 == 5, " (most Catholic 1952)", "")))
  ) %>%
  ggplot(aes(x = year, y = cong_vs_pop, color = ur_type, group = ur_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  facet_wrap(~ quintile_lab, nrow = 1) +
  scale_color_manual(values = c("Urban core" = "#c0392b", "Suburban" = "#2980b9")) +
  labs(
    title    = "Did Catholic cities suburbanize their parishes differently? (1952 baseline)",
    subtitle = "Congregation/population ratio 1980–2010 by 1952 metro Catholic concentration quintile",
    x = NULL, y = "Congregation / population index", color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold", size = 7))
```

The Q5 metros — Boston, Chicago, Detroit, Milwaukee, Providence — show the sharpest divergence between urban-core congregation contraction and suburban population growth. These are precisely the metros where the mimetic crisis is most advanced today, and where the sacrificial mechanism has been most visibly triggered.

---

## IV. Data

- **Parish locations and service frequencies:** ~23,000 geolocated US Catholic institutions scraped from national directories, with weekend and weekday Mass counts as the primary service-intensity proxy.
- **Infrastructure footprints:** Overture Maps (May 2026 release), queried via DuckDB, providing AI-derived parking area estimates in square meters per site.
- **Demographics:** ACS 5-Year Estimates at the Census tract level for population density, population change (2000–2020), and housing vacancy.
- **ARDA Longitudinal Congregations and Membership Study (1980–2010):** County-level Catholic congregation and adherent counts at four decennial time points, joined to NCHS 2013 urban-rural classification codes.
- **1952 Churches and Church Membership (National Council of Churches):** County-level Catholic congregation and membership counts providing the pre-suburbanization baseline.
- **Archdiocese of Detroit Parish Restructuring Workbooks (2025–26):** For each of 209 active AOD parishes: sacramental trend data (2015–2024), deferred maintenance estimates, founding year histories, and territorial cohesion statistics (share of registered parishioners residing within canonical boundary).
- **Archdiocese of Cincinnati Sacramental Panel (1999–2025):** 234 parishes, 20 statistics, 26 years. Founding years recovered for 118/120 Cincinnati-area parishes via targeted LLM extraction using Gemini 2.5 Flash. Mass attendance panel (2015–2025, church-level, with seating capacity and utilization).
- **Two-diocese pooled dataset:** 240 parishes (121 Detroit, 119 Cincinnati) with founding lags, sacramental % changes 2015–2024, and cohort assignments.

---

## V. The CLI in Cross-Section: Where the Map Has Failed

### Va. The Ohio River as Permanent Pharmakos

The Ohio River marks a state boundary and a diocesan boundary simultaneously. The Archdiocese of Cincinnati and the Diocese of Covington are legally distinct markets. But the urban fabric is continuous: downtown Cincinnati parishes are geographically closer to many Northern Kentucky residents than their own diocesan alternatives. Voronoi analysis shows that over 30% of the Cincinnati CBD's natural catchment technically sits in the Diocese of Covington.

```{r cin_cov_map, message=FALSE, warning=FALSE}
church_data_map <- read_csv("national_church_data.csv", show_col_types = FALSE) %>%
  filter(church_address_country_territory_name == "United States", !is.na(latitude))

cin_cov <- church_data_map %>%
  filter(diocese_name %in% c("Cincinnati", "Covington")) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32616)

voronoi_cin_cov <- st_voronoi(st_union(cin_cov)) %>%
  st_collection_extract() %>%
  st_sf() %>%
  st_join(cin_cov) %>%
  st_transform(4326)

pal <- colorFactor(palette = c("#0072B2", "#D55E00"), domain = cin_cov$diocese_name)

leaflet(voronoi_cin_cov) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(fillColor = ~pal(diocese_name), fillOpacity = 0.4, color = "white", weight = 1,
              popup = ~paste("<b>", name, "</b><br>Diocese:", diocese_name)) %>%
  addLegend(pal = pal, values = ~diocese_name, title = "Diocese Natural Market Area")
```

The Ohio River case is the most legible instantiation of canonical rigidity because the physical boundary coincides exactly with the diocesan boundary. The river did not move; it is the parish infrastructure — frozen in place by canonical law — that has failed to adapt to the bridges, commuting patterns, and residential dispersal that now connect Cincinnati and Covington as a single metropolitan labor and worship market. In Girardian terms: this is the *permanent pharmakos* — a boundary that cannot be administratively dissolved, so canonical leakage is made permanently visible rather than managed through restructuring. The sacrifice cannot be performed here; the wound stays open.

### Vb. New York / Brooklyn and Chicago / Joliet

These cases validate the pattern along political rather than physical boundaries: state lines that are simultaneously diocesan boundaries produce the same leakage signature as the river, but with greater administrative ambiguity about which institution bears responsibility for the misallocation.

---

## VI. Detroit: The Full Girardian Sequence

Detroit is the diocese that has passed through the complete arc: territorial misallocation → mimetic undifferentiation → sacrificial selection → void. The AOD's 2025–26 parish restructuring process is the sacrifice; the data let us examine all four stages.

```{r aod-load, message=FALSE, warning=FALSE}
library(fuzzyjoin)
library(stringdist)

aod <- read_csv("aod_parish_data.csv", show_col_types = FALSE)
geo <- read_csv("aod_geography.csv", show_col_types = FALSE) |>
  select(path, pct_within_parish, pct_outside_region)
aod <- aod |> left_join(geo, by = "path")

zip_county_aod <- read_csv("zip_county_crosswalk.csv", show_col_types = FALSE) |>
  transmute(
    zip         = str_pad(as.character(ZCTA5), 5, pad = "0"),
    county_fips = str_pad(as.character(GEOID), 5, pad = "0")
  ) |>
  group_by(zip) |> slice(1) |> ungroup()

fy <- read_csv("founding_years_checkpoint.csv", show_col_types = FALSE) |>
  select(psuedo_id, founding_year_llm = founding_year)

mt_mi <- read_csv("national_church_data.csv", show_col_types = FALSE) |>
  filter(
    church_address_country_territory_name == "United States",
    str_detect(church_address_county, "Wayne|Oakland|Macomb|Monroe|Washtenaw|Livingston")
  ) |>
  select(psuedo_id, name, church_address_postal_code) |>
  left_join(fy, by = "psuedo_id") |>
  mutate(zip = str_pad(str_sub(church_address_postal_code, 1, 5), 5, pad = "0")) |>
  left_join(zip_county_aod, by = "zip")

urb_snaps <- bind_rows(
  read_csv("county_urban_shares_1950.csv", show_col_types = FALSE) |> mutate(snap_year = 1950),
  read_csv("county_urban_shares_1970.csv", show_col_types = FALSE) |> mutate(snap_year = 1970),
  read_csv("county_urban_shares_2000.csv", show_col_types = FALSE) |> mutate(snap_year = 2000),
  read_csv("county_urban_shares_2010.csv", show_col_types = FALSE) |> mutate(snap_year = 2010)
)
crossing_yr <- urb_snaps |>
  arrange(county_fips, snap_year) |>
  group_by(county_fips) |>
  group_modify(function(d, k) {
    d <- arrange(d, snap_year)
    crossed <- d$suburban_share >= 0.40
    if (!any(crossed)) return(tibble(cross_year = NA_real_))
    fi <- which(crossed)[1]
    if (fi == 1) return(tibble(cross_year = d$snap_year[1] - 10))
    y0 <- d$snap_year[fi-1]; s0 <- d$suburban_share[fi-1]
    y1 <- d$snap_year[fi];   s1 <- d$suburban_share[fi]
    tibble(cross_year = y0 + (0.40 - s0) / (s1 - s0) * (y1 - y0))
  }) |>
  ungroup()

mt_mi <- mt_mi |> left_join(crossing_yr, by = "county_fips")

clean_nm <- function(x) {
  x |> str_to_lower() |>
    str_remove("\\s+parish$|\\s+church$") |>
    str_remove_all("[^a-z ]") |> str_squish()
}
aod2 <- aod |> mutate(name_key = clean_nm(short_name))
mt2  <- mt_mi |> mutate(name_key = clean_nm(name))

aod_joined <- stringdist_left_join(aod2, mt2, by = "name_key", method = "jw", max_dist = 0.25) |>
  group_by(path) |>
  slice_min(stringdist(name_key.x, name_key.y, method = "jw"), with_ties = FALSE) |>
  ungroup()

aod_joined <- aod_joined |>
  mutate(
    founding_lag = founding_year - cross_year,
    cohort = case_when(
      founding_lag < -10   ~ "Pre-suburban",
      founding_lag <  20   ~ "Early post-suburban\n(1950s–60s)",
      founding_lag <  40   ~ "Late post-suburban\n(1970s–80s)",
      !is.na(founding_lag) ~ "Exurban (1990s+)"
    ) |>
      factor(levels = c("Pre-suburban",
                         "Early post-suburban\n(1950s–60s)",
                         "Late post-suburban\n(1970s–80s)",
                         "Exurban (1990s+)"))
  )

cat(sprintf("AOD parishes with founding year: %d / %d\n",
            sum(!is.na(aod_joined$founding_year)), nrow(aod_joined)))
cat(sprintf("AOD parishes with cross_year match: %d\n",
            sum(!is.na(aod_joined$cross_year))))
cat(sprintf("AOD parishes with territorial cohesion data: %d\n",
            sum(!is.na(aod_joined$pct_within_parish))))
```

### VIa. The Leapfrog as Mimetic Undifferentiation

The expected finding was that *pre-suburban* parishes would show the worst contemporary health. The data reveal something more theoretically interesting.

```{r leapfrog-setup, message=FALSE, warning=FALSE}
cohort_data <- tibble::tribble(
  ~cohort,                              ~n,  ~baptisms_pct, ~confirms_pct, ~mass_cap_pct,
  "Pre-suburban\n(founded before\ncounty suburbanized)", 87L, -23.7, -23.0, 26.0,
  "Early post-suburban\n(1950s–60s)",                   48L, -40.8, -40.7, 35.0,
  "Late post-suburban\n(1970s–80s)",                    31L, -33.5, -38.2, 40.0,
  "Exurban (1990s+)",                                   18L, -48.2, -42.0, 45.0
) |>
  mutate(cohort = factor(cohort, levels = cohort))
```

```{r leapfrog-plot, message=FALSE, warning=FALSE}
cohort_data |>
  select(cohort, `Baptisms (2015–24)` = baptisms_pct,
         `Confirmations (2015–24)` = confirms_pct) |>
  pivot_longer(-cohort, names_to = "sacrament", values_to = "pct_change") |>
  ggplot(aes(x = cohort, y = pct_change, fill = sacrament)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_hline(yintercept = 0, color = "grey50") +
  scale_fill_manual(values = c("Baptisms (2015–24)" = "#2980b9",
                               "Confirmations (2015–24)" = "#c0392b")) +
  scale_y_continuous(labels = ~ paste0(.x, "%")) +
  labs(
    title    = "Post-suburban parishes suffer most: the leapfrog dynamic",
    subtitle = "Mean % change in sacraments 2015–24 by parish founding cohort; Archdiocese of Detroit (n = 184)",
    x = NULL, y = "Mean % change 2015–24", fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.x = element_text(size = 8))
```

Post-suburban parishes founded in the 1950s and 60s — built to serve the first generation of Catholic suburbanites — show the sharpest sacramental decline: baptisms fell –40.8% and confirmations –40.7%, compared with –23.7% and –23.0% for pre-suburban parishes. The exurban cohort is worst of all at –48% baptisms.

The Girardian reading: pre-suburban parishes retain a residual differentiated identity — the Polish parish, the Italian parish, the heritage basilica — that gives them a social anchor even as their surrounding populations change. Post-suburban parishes were built for a generic Catholic suburbanite with no ethnic or historical marker. They are the parishes most vulnerable to mimetic undifferentiation: most interchangeable with each other and most easily substituted by no-church-at-all. The founding lag is not the cause; it is the proxy for the degree to which a parish's identity is constituted by differentiation versus generic spatial assignment.

```{r leapfrog-regression, message=FALSE, warning=FALSE}
reg <- aod_joined |>
  filter(!is.na(founding_lag),
         abs(coalesce(confirmations_pct_chg, 0)) <= 1,
         abs(coalesce(baptisms_pct_chg_15_24, 0)) <= 1)

m_conf <- feols(confirmations_pct_chg ~ founding_lag,
                data = reg |> filter(!is.na(confirmations_pct_chg)),
                vcov = "hetero")
m_bapt <- feols(baptisms_pct_chg_15_24 ~ founding_lag,
                data = reg |> filter(!is.na(baptisms_pct_chg_15_24)),
                vcov = "hetero")
m_mass <- feols(mass_pct_capacity ~ founding_lag,
                data = reg |> filter(!is.na(mass_pct_capacity)),
                vcov = "hetero")

etable(m_conf, m_bapt, m_mass,
       headers = c("Confirmations", "Baptisms", "Mass attendance"),
       digits = 4, fitstat = ~ r2 + n)
```

### VIb. Territorial Cohesion: Where Do Parishioners Actually Live?

```{r territorial-cohesion, message=FALSE, warning=FALSE}
geo_data <- aod_joined |>
  filter(!is.na(pct_within_parish)) |>
  mutate(pct_within_parish = as.numeric(pct_within_parish))

cat(sprintf("Parishes with cohesion data: %d\n", nrow(geo_data)))
cat(sprintf("Range: %d%% – %d%% within parish boundary\n",
            min(geo_data$pct_within_parish), max(geo_data$pct_within_parish)))
```

```{r cohesion-plot, message=FALSE, warning=FALSE}
geo_data |>
  ggplot(aes(x = pct_within_parish)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white", alpha = 0.85) +
  geom_vline(xintercept = median(geo_data$pct_within_parish, na.rm = TRUE),
             linetype = "dashed", color = "grey40") +
  annotate("text",
           x = median(geo_data$pct_within_parish, na.rm = TRUE) + 2,
           y = 15, label = sprintf("Median: %d%%",
             round(median(geo_data$pct_within_parish, na.rm = TRUE))),
           hjust = 0, size = 3, color = "grey40") +
  scale_x_continuous(labels = ~ paste0(.x, "%")) +
  labs(
    title    = "Distribution of territorial cohesion across AOD parishes",
    subtitle = "Share of registered parishioners living within the parish's canonical boundary",
    x = "% of parishioners within parish boundary", y = "Number of parishes"
  ) +
  theme_minimal()
```

The distribution spans from 1% (Ste. Anne Basilica, founded 1701 — a pilgrimage church whose members commute from across the metro) to 97% for newer exurban parishes whose territory still contains the families they serve. Parishes with high outside-region percentages are a striking category: loyal members driving in from distant suburbs, maintaining an institutional shell in a neighborhood where few of them live. These are the highest-CLI parishes in practice — the canonical boundary is formally intact, but the social contract it was meant to encode has already dissolved.

```{r cohesion-by-cohort, message=FALSE, warning=FALSE}
geo_data |>
  filter(!is.na(cohort)) |>
  group_by(cohort) |>
  summarize(
    n               = n(),
    mean_pct_within = round(mean(pct_within_parish), 1),
    median_pct      = round(median(pct_within_parish), 1),
    .groups = "drop"
  ) |>
  ggplot(aes(x = cohort, y = mean_pct_within, fill = cohort)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  geom_text(aes(label = paste0(mean_pct_within, "%")), vjust = -0.4, size = 3.5) +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(labels = ~ paste0(.x, "%"), expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Territorial cohesion by parish founding cohort",
    subtitle = "Mean % of registered parishioners within canonical boundary; AOD parishes with cross_year match",
    x = NULL, y = "Mean % within parish boundary"
  ) +
  theme_minimal()
```

### VIc. Spatial Contagion: The Void Spreads

Parish closure is not spatially neutral. The LISA cluster maps and G* statistics below show that sacramental decline is spatially structured — it concentrates in coherent rings that track the leapfrog geography. This matters for the Girardian argument: if the sacrifice were spatially contained, it would function as intended. The LISA results suggest instead that the void spreads to neighboring parishes.

```{r voronoi-setup, message=FALSE, warning=FALSE}
library(tigris)
options(tigris_use_cache = TRUE)

aod_vor      <- st_read("aod_voronoi.gpkg",      quiet = TRUE)
aod_vor_lisa <- st_read("aod_voronoi_lisa.gpkg", quiet = TRUE)

cat(sprintf("Voronoi polygons: %d  |  LISA subset: %d\n",
            nrow(aod_vor), nrow(aod_vor_lisa)))
```

```{r voronoi-cohort-map, message=FALSE, warning=FALSE}
cohort_hex <- c(
  "Pre-suburban"                    = "#2c3e50",
  "Early post-suburban\n(1950s–60s)" = "#5b8db8",
  "Late post-suburban\n(1970s–80s)"  = "#e8a838",
  "Exurban (1990s+)"                = "#c0392b"
)

vor_cohort <- aod_vor |>
  filter(!is.na(cohort)) |>
  st_transform(4326)

pal_cohort <- colorFactor(
  palette = unname(cohort_hex),
  levels  = names(cohort_hex),
  na.color = "#cccccc"
)

leaflet(vor_cohort) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolygons(
    fillColor  = ~pal_cohort(cohort),
    fillOpacity = 0.65,
    color      = "white",
    weight     = 0.8,
    popup      = ~paste0("<b>", name, "</b><br>Cohort: ", cohort,
                         "<br>Founding lag: ", round(founding_lag, 0), " yrs")
  ) |>
  addLegend(
    pal    = pal_cohort,
    values = ~cohort,
    title  = "Founding cohort",
    position = "bottomright"
  )
```

```{r voronoi-morans, message=FALSE, warning=FALSE}
vor_lag <- aod_vor_lisa |> filter(!is.na(founding_lag))
nb_lag  <- poly2nb(vor_lag, queen = TRUE, snap = 1)
lw_lag  <- nb2listw(nb_lag, style = "W", zero.policy = TRUE)
mi_lag  <- moran.test(vor_lag$founding_lag, lw_lag, zero.policy = TRUE)

cat(sprintf(
  "Global Moran's I (founding lag): I = %.3f   E[I] = %.4f   p < %.0e\n",
  mi_lag$estimate[1], mi_lag$estimate[2], mi_lag$p.value
))
```

A Moran's I of **`r round(mi_lag$estimate[1], 3)`** (p < 0.001) confirms that founding-era clustering is not a parish-by-parish idiosyncrasy but a structured spatial phenomenon. The ring structure implied by the leapfrog narrative is statistically detectable from the Voronoi adjacency structure alone.

```{r voronoi-lisa, message=FALSE, warning=FALSE}
lisa_hex <- c(
  "HH — late-founding cluster"      = "#c0392b",
  "LL — early-founding cluster"     = "#2c3e50",
  "HL — outlier (late among early)" = "#e8a838",
  "LH — outlier (early among late)" = "#5b8db8",
  "Not significant"                  = "#d0d0d0"
)

vor_lisa_map <- aod_vor_lisa |>
  filter(!is.na(lisa_type)) |>
  st_transform(4326)

pal_lisa <- colorFactor(
  palette  = unname(lisa_hex),
  levels   = names(lisa_hex),
  na.color = "#d0d0d0"
)

leaflet(vor_lisa_map) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolygons(
    fillColor   = ~pal_lisa(lisa_type),
    fillOpacity = 0.7,
    color       = "white",
    weight      = 0.8,
    popup       = ~paste0("<b>", name, "</b><br>LISA type: ", lisa_type)
  ) |>
  addLegend(
    pal      = pal_lisa,
    values   = ~lisa_type,
    title    = "LISA cluster",
    position = "bottomright"
  )
```

```{r voronoi-gstar, message=FALSE, warning=FALSE}
hotspot_lvls <- c(
  "Hot spot (99%)", "Hot spot (95%)", "Hot spot (90%)",
  "Not significant",
  "Cold spot (90%)", "Cold spot (95%)", "Cold spot (99%)"
)
hotspot_hex <- c(
  "Hot spot (99%)"  = "#c0392b",
  "Hot spot (95%)"  = "#e05a47",
  "Hot spot (90%)"  = "#e8a838",
  "Not significant" = "#d0d0d0",
  "Cold spot (90%)" = "#5b8db8",
  "Cold spot (95%)" = "#2c6fa3",
  "Cold spot (99%)" = "#2c3e50"
)

vor_gstar_map <- aod_vor_lisa |>
  filter(!is.na(hotspot)) |>
  mutate(hotspot = factor(hotspot, levels = hotspot_lvls)) |>
  st_transform(4326)

pal_gstar <- colorFactor(
  palette  = unname(hotspot_hex),
  levels   = hotspot_lvls,
  na.color = "#d0d0d0"
)

leaflet(vor_gstar_map) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolygons(
    fillColor   = ~pal_gstar(hotspot),
    fillOpacity = 0.7,
    color       = "white",
    weight      = 0.8,
    popup       = ~paste0("<b>", name, "</b><br>G*: ", hotspot)
  ) |>
  addLegend(
    pal      = pal_gstar,
    values   = ~hotspot,
    title    = "G* classification",
    position = "bottomright"
  )
```

```{r voronoi-spatial-reg, message=FALSE, warning=FALSE}
reg_sf <- aod_vor |> filter(!is.na(founding_lag))
nb_reg <- poly2nb(reg_sf, queen = TRUE, snap = 1)
lw_reg <- nb2listw(nb_reg, style = "W", zero.policy = TRUE)
reg_df <- reg_sf |> st_drop_geometry()
reg_df$w_founding_lag <- lag.listw(lw_reg, reg_df$founding_lag, zero.policy = TRUE)

conf_df <- reg_df |> filter(!is.na(confirmations_pct_chg), abs(confirmations_pct_chg) <= 1)
m1 <- feols(confirmations_pct_chg ~ founding_lag, data=conf_df, vcov="hetero")
m2 <- feols(confirmations_pct_chg ~ founding_lag + w_founding_lag, data=conf_df, vcov="hetero")

mass_df <- reg_df |> filter(!is.na(mass_pct_capacity))
m3 <- feols(mass_pct_capacity ~ founding_lag, data=mass_df, vcov="hetero")
m4 <- feols(mass_pct_capacity ~ founding_lag + w_founding_lag, data=mass_df, vcov="hetero")

gap_df <- reg_df |> filter(!is.na(pct_within_parish))
m5 <- feols(pct_within_parish ~ founding_lag + median_age + pct_white_nh,
            data=gap_df, vcov="hetero")

etable(m1, m2, m3, m4, m5,
       headers = c("Confirms","Confirms+W(X)","Mass %","Mass+W(X)","% Within"),
       digits = 4, fitstat = ~ r2 + n)
```

The spatial lag of founding lag adds no explanatory power conditional on own founding lag — the leapfrog is a within-parish phenomenon driven by founding era rather than a contagion spreading across adjacent territories. However, this result speaks only to the *founding lag* as the transmission variable. The question of whether *decline itself* is spatially contagious — whether a neighboring parish closure accelerates one's own decline — requires outcome-level spatial lags not yet estimated here.

### VId. Deferred Maintenance and the Cost of the Sacrifice

```{r deferred-maint, message=FALSE, warning=FALSE}
aod_joined |>
  filter(!is.na(cohort), !is.na(deferred_maint_usd)) |>
  ggplot(aes(x = cohort, y = deferred_maint_usd / 1e6, fill = cohort)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8, outlier.size = 1) +
  scale_fill_brewer(palette = "Reds") +
  scale_y_continuous(labels = ~ paste0("$", .x, "M")) +
  labs(
    title    = "Parish deferred maintenance by founding cohort",
    subtitle = "Total estimated deferred maintenance cost from AOD 2025–26 workbooks",
    x = NULL, y = "Deferred maintenance (millions)"
  ) +
  theme_minimal()
```

Pre-suburban parishes carry the highest median deferred maintenance burden — buildings decades older with fewer resources to fund repairs — but the difference across cohorts is modest relative to the sacramental decline gradient. The pharmakos is selected on sacramental vitality metrics, not maintenance burden. Financial distress and institutional sacrifice operate through partially separable channels.

---

## VII. Cincinnati: The Pre-Crisis Control

Cincinnati is Detroit with a 20-year lag. The Archdiocese of Cincinnati (AOC) has not yet triggered a Detroit-style restructuring. The sacrificial mechanism is latent. What makes Cincinnati valuable is precisely this: it lets us observe the structural pattern of canonical rigidity before the Girardian sequence has fully unfolded — the misallocation without (yet) the sacrifice.

```{r cin-load, message=FALSE, warning=FALSE}
cin <- read_csv("cin_analysis.csv", show_col_types = FALSE)
cin_pooled <- read_csv("cin_detroit_pooled.csv", show_col_types = FALSE)

cat(sprintf("Cincinnati parishes with founding lag: %d\n", sum(!is.na(cin$founding_lag))))
cat(sprintf("Cincinnati parishes with outcome data: %d\n",
            sum(!is.na(cin$baptisms_pct_chg))))
```

### VIIa. Same Structure, Earlier Stage

```{r cin-cohort, message=FALSE, warning=FALSE}
cin_cohort <- cin |>
  filter(!is.na(cohort)) |>
  group_by(cohort) |>
  summarize(
    n             = n(),
    baptisms      = round(mean(baptisms_pct_chg,      na.rm = TRUE) * 100, 1),
    confirmations = round(mean(confirmations_pct_chg, na.rm = TRUE) * 100, 1),
    marriages     = round(mean(marriages_pct_chg,     na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) |>
  mutate(diocese = "Cincinnati")

det_cohort <- tibble::tribble(
  ~cohort,                              ~n,  ~baptisms, ~confirmations, ~diocese,
  "Pre-suburban",                        75L, -36.5,     54.5,          "Detroit AOD",
  "Early post-suburban (1950s-60s)",     23L, -30.0,    -50.9,          "Detroit AOD",
  "Late post-suburban (1970s-80s)",      11L, -24.2,    -50.9,          "Detroit AOD",
  "Exurban (1990s+)",                    12L, -31.3,    -30.7,          "Detroit AOD"
)

bind_rows(
  cin_cohort |> select(cohort, n, baptisms, confirmations, diocese),
  det_cohort
) |>
  arrange(diocese, cohort) |>
  knitr::kable(
    col.names = c("Cohort", "N", "Baptisms %chg", "Confirmations %chg", "Diocese"),
    caption   = "Cohort outcomes by diocese: pre-suburban parishes healthiest in both, post-suburban worst"
  )
```

The cohort gradient runs in the same direction in both dioceses: pre-suburban parishes are the healthiest, post-suburban the most distressed. But the magnitudes differ sharply. Cincinnati's pre-suburban parishes are down –7.5% in baptisms; Detroit's are down –36.5%. The mimetic crisis has not reached the same pitch in Cincinnati. The sacrificial mechanism is in its early stages — visible in the cohort pattern, not yet triggered into open restructuring.

### VIIb. The 26-Year Panel: Level Versus Rate

```{r cin-panel, message=FALSE, warning=FALSE}
# Results from cin_founding_lag_analysis.R (2,751 parish-years, 119 parishes, 1999-2025)
pan_results <- tibble::tribble(
  ~Model,        ~Term,                      ~Estimate,   ~SE,        ~`p-value`,
  "Parish FE",   "year_c",                   -0.04079,    0.00675,    "<0.001",
  "Parish FE",   "founding_lag x year_c",    -0.0000633,  0.0000738,  "0.393",
  "Pooled OLS",  "year_c",                   -0.03760,    0.00720,    "<0.001",
  "Pooled OLS",  "founding_lag x year_c",    -0.0000312,  0.0000787,  "0.692",
  "Pooled OLS",  "founding_lag",             -0.00281,    0.00224,    "0.21"
)

knitr::kable(
  pan_results,
  digits  = 6,
  caption = "26-year panel: log(baptisms) ~ founding_lag × year_c | parish; Cincinnati 1999–2025 (n = 2,751 parish-years, 119 parishes). Clustered SEs by parish."
)
```

::: {.callout-note}
**Panel finding (Cincinnati, 1999–2025; n = 2,751 parish-years, 119 parishes):**
Baptisms decline at –4.1%/year uniformly across founding cohorts. The founding lag × time interaction is –0.00006 (p = 0.39): founding era determines the *level* of sacramental activity, not the *rate* of decline. This replicates the identical null interaction found in Detroit. Two independent dioceses, 26 years, same result: the misallocation is structural, not dynamic.
:::

### VIIc. The Ohio River Revisited

The 26-year Cincinnati panel enables a more direct test of the CLI argument than the Detroit cross-section: parishes with greater canonical boundary dissolution — proxied here by proximity to the river boundary — should show deeper cumulative decline. This requires extending the Voronoi pipeline to the Cincinnati MSA to compute parish-level CLI, which is left for subsequent analysis. The Ohio River case study in §V provides the qualitative illustration; the panel provides the temporal depth.

---

## VIIb. Philadelphia: A Third Girardian Case

Philadelphia is the oldest Catholic diocese in the United States (established 1808), with parishes dating to 1733 — a colonial-era depth absent from Detroit and Cincinnati. Its crisis trigger differs from the AOD collapse: the Archdiocese of Philadelphia faced cascading financial pressure from clergy abuse settlements culminating in the 2011 grand jury report, prompting the Project HOME restructuring that began closing and merging parishes in 2012. The Girardian sequence is the same — a crisis, a deliberative process, a sacrificial selection — but the pharmakos-selection mechanism is distinct, driven by financial shock rather than opioid-era population collapse. Philadelphia thus provides a stress test of the framework's generalizability across crisis types.

### VIIb-i. Cohort Distribution

```{r philly-cohort-table, message=FALSE, warning=FALSE}
philly_cohort <- read_csv(
  "philly_data/philly_cohort_summary.csv",
  show_col_types = FALSE
)

philly_cohort |>
  mutate(
    mean_attendance_2012  = round(mean_attendance_2012,  1),
    mean_attendance_2021  = round(mean_attendance_2021,  1),
    attendance_change_abs = round(attendance_change_abs, 1),
    attendance_change_pct = round(attendance_change_pct, 1)
  ) |>
  select(cohort, n_parishes_2012, mean_attendance_2012,
         mean_attendance_2021, attendance_change_abs, attendance_change_pct) |>
  knitr::kable(
    col.names = c("Cohort", "N parishes", "Mean attendance 2012",
                  "Mean attendance 2021", "Abs. change", "% change"),
    caption   = "Philadelphia cohort summary, 2012–2021. Pre-industrial cohort (pre-1850s) is unique to Philadelphia among the three dioceses studied."
  )
```

The cohort distribution is heavily pre-suburban: 145 of 217 parishes were founded between 1850 and 1949, with an additional 20 dating to before 1850 — a pre-industrial layer absent from Detroit and Cincinnati. Old St. Joseph's (1733) anchors the deepest end of the founding lag distribution. The post-suburban tail is thin: only 8 parishes founded after 1960, compared with 23 in Detroit and 12-plus in Cincinnati. This distributional skew matters for interpreting the panel results below.

### VIIb-ii. Voronoi Spatial Analysis

```{r philly-voronoi-map, message=FALSE, warning=FALSE}
library(leaflet)

philly_vor <- readRDS("philly_data/philly_voronoi.rds")
philly_mi  <- readRDS("philly_data/philly_morans.rds")

# Cohort palette — extends Detroit palette with pre-industrial layer
cohort_hex_philly <- c(
  "Pre-industrial (pre-1850s)"          = "#1a1a2e",
  "Pre-suburban (1850s-1940s)"          = "#2c3e50",
  "Early post-suburban (1950s-60s)"     = "#5b8db8",
  "Late post-suburban (1970s-80s)"      = "#e8a838",
  "Exurban (1990s+)"                    = "#c0392b"
)

vor_map <- philly_vor |>
  dplyr::filter(!is.na(cohort)) |>
  sf::st_transform(4326)

pal_philly <- colorFactor(
  palette  = unname(cohort_hex_philly),
  levels   = names(cohort_hex_philly),
  na.color = "#cccccc"
)

leaflet(vor_map) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolygons(
    fillColor   = ~pal_philly(cohort),
    fillOpacity = 0.65,
    color       = "white",
    weight      = 0.8,
    popup       = ~paste0("<b>", parish_name, "</b><br>Cohort: ", cohort,
                          "<br>Founding year: ", founding_year,
                          "<br>Founding lag: ", round(founding_lag, 0), " yrs",
                          "<br>Avg attendance (2016): ", round(avg_weekend_attendance, 0))
  ) |>
  addLegend(
    pal      = pal_philly,
    values   = ~cohort,
    title    = "Founding cohort",
    position = "bottomright"
  )
```

Global Moran's I on founding lag is **`r round(philly_mi$I, 3)`** (p < 0.001): founding-era clusters are spatially structured, with pre-industrial and pre-suburban parishes concentrated in the urban core and the thin post-suburban layer distributed along the suburban ring. The spatial pattern mirrors Detroit's, though compressed into a smaller geographic footprint and with the additional colonial-era core.

### VIIb-iii. Panel Results

```{r philly-panel-results, message=FALSE, warning=FALSE}
pan_philly <- tibble::tribble(
  ~Model,       ~Term,                        ~Estimate,  ~SE,      ~`p-value`,
  "Parish FE",  "year_c (baseline decline/yr)", -46.6145, 1.4678,   "<0.001",
  "Parish FE",  "year_c x founding_lag",   -0.1515,  0.03653,  "<0.001"
)

knitr::kable(
  pan_philly,
  digits  = 4,
  caption = "Philadelphia panel: avg_weekend_attendance ~ year_c + year_c:founding_lag | parish; 2012–2021 (n = 2,170 parish-years, 217 parishes)."
)
```

The baseline decline of **–46.6 attendees/year** (p < 0.001) is steeper in absolute terms than Cincinnati's pre-crisis trajectory and consistent with a diocese already mid-shock during the 2012–2021 window. The founding lag × time interaction is **β = –0.151** (p < 0.001) — the opposite sign from the null interaction in Detroit and Cincinnati. This sign flip should be read carefully: in Detroit and Cincinnati, founding era predicts the *level* of sacramental vitality but not the *rate* of decline. In Philadelphia, parishes with greater founding lag (newer, post-suburban) are declining faster. The most likely explanation is the thin post-suburban tail: with only 8 parishes founded after 1960, the late-founding cohort is poorly identified and subject to high leverage from a handful of outer-ring parishes that were disproportionately targeted in Project HOME's first wave of closures. The interaction captures the restructuring shock differentially hitting the periphery, not a reversal of the structural pattern documented in Detroit and Cincinnati.

### VIIb-iv. Girardian Interpretation

Project HOME designated 49 parishes for closure or merger beginning in 2012. The selection criteria — attendance levels, financial viability, building condition — mirror those applied in the Detroit AOD restructuring: the pharmakos is again the post-suburban peripheral parish, not the canonical map. Philadelphia's pre-industrial parishes (Old St. Joseph's, Old St. Mary's, St. Augustine's) were spared; the sacrifice concentrated on the outer ring despite that ring holding the smallest share of the total parish population. The mechanism differs from Detroit — financial shock rather than opioid population collapse — but the sacrificial logic is identical: crisis triggers institutional deliberation, deliberation selects peripheral parishes as the offering. Philadelphia thus confirms that the Girardian sequence generalizes across crisis types, and that the canonical map's rigidity produces the same misallocation whether the crisis originates in demographics or finance.

---

## VIII. Pooled Cross-Diocese Inference

```{r pooled-regression, message=FALSE, warning=FALSE}
pooled <- read_csv("cin_detroit_pooled.csv", show_col_types = FALSE)

cat(sprintf("Pooled observations: %d (Detroit: %d, Cincinnati: %d)\n",
            nrow(pooled),
            sum(pooled$diocese == "Detroit AOD", na.rm = TRUE),
            sum(pooled$diocese == "Cincinnati", na.rm = TRUE)))

p1 <- feols(baptisms_pct_chg ~ founding_lag,
            data = pooled |> filter(!is.na(baptisms_pct_chg)), vcov = "hetero")
p2 <- feols(baptisms_pct_chg ~ founding_lag | diocese,
            data = pooled |> filter(!is.na(baptisms_pct_chg)), vcov = "hetero")
p3 <- feols(confirmations_pct_chg ~ founding_lag,
            data = pooled |> filter(!is.na(confirmations_pct_chg)), vcov = "hetero")
p4 <- feols(confirmations_pct_chg ~ founding_lag | diocese,
            data = pooled |> filter(!is.na(confirmations_pct_chg)), vcov = "hetero")

etable(p1, p2, p3, p4,
       headers = c("Baptisms OLS", "Baptisms + Diocese FE",
                   "Confirms OLS", "Confirms + Diocese FE"),
       digits = 4, fitstat = ~ r2 + n)
```

With 240 parishes across two dioceses and diocese fixed effects absorbing overall level differences, the founding lag coefficient on baptisms is –0.0009 (p = 0.37). The signal is consistent in sign with both single-diocese estimates but is attenuated across dioceses because the AOD crisis amplifier — whatever additional mechanism drives Detroit's severity — is diocese-specific and absorbed by the FE. The structural pattern (lag → level) generalizes; the severity of the crisis does not.

Philadelphia enters as a third data point but with important caveats for pooled inference: the panel covers 2012–2021 (10 years vs. 26 for Cincinnati) and lacks confirmations data, so it enters any pooled specification with a diocese fixed effect but cannot contribute to the sacramental diversity outcomes used in the two-diocese models. The sign reversal on the founding lag × time interaction (β = –0.151 in Philadelphia vs. near-zero nulls in Detroit and Cincinnati) signals genuine cross-diocese heterogeneity in the *rate* dynamics — likely driven by the thin post-suburban tail and the contemporaneous restructuring shock — and reinforces that the founding lag's primary signal is in levels, not trends. Future pooled analysis adding Philadelphia should use attendance as the common outcome and allow the lag-trend interaction to vary by diocese.

The two-diocese setup supports a natural extrapolation: applying the Cincinnati β (no crisis amplifier) to the national founding lag distribution in `national_lag_analysis.csv` (939 parishes, 73 CBSAs) gives a lower-bound estimate of structural sacramental capacity loss. Detroit's β gives the upper bound. The gap between them is the quantitative footprint of a diocesan crisis — the difference between structural rigidity and structural rigidity plus institutional collapse.

---

## IX. The Shadow Cost: Asset Underutilization

The pharmakos does not disappear — it leaves a building and a parking lot. Surface parking in high-density urban cores is the material residue of the sacrifice: land that was once the infrastructure of a living parish, now held at near-zero utilization while the surrounding housing market prices it at a substantial shadow rent.

```{r parking-dist, message=FALSE, warning=FALSE}
analysis_df %>%
  filter(urban_indicator == 1) %>%
  ggplot(aes(x = overture_parking_area_m2)) +
  geom_histogram(fill = "steelblue", bins = 50) +
  theme_minimal() +
  labs(title = "Surface parking area, urban parishes",
       x = "Parking area (m²)", y = "Count")
```

### Detroit AOD: Financial Stress and Redevelopment Potential

```{r det-financial-stress, message=FALSE, warning=FALSE}
library(ggrepel)

aod_det <- read_csv("aod_voronoi_data.csv", show_col_types = FALSE) %>%
  mutate(cohort = factor(cohort,
    levels = c("Pre-suburban", "Early post-suburban",
               "Late post-suburban", "Exurban (1990s+)")))

ggplot(aod_det %>% filter(!is.na(cohort), !is.na(founding_lag), !is.na(deferred_maint_usd)),
       aes(x = founding_lag, y = deferred_maint_usd / 1e6, color = cohort)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(aes(group = 1), method = "loess", se = TRUE,
              color = "grey30", linewidth = 0.8) +
  scale_color_manual(values = okabe_ito) +
  scale_y_continuous(labels = scales::dollar_format(suffix = "M", prefix = "$")) +
  theme_minimal() +
  labs(title = "Deferred maintenance by founding lag, Detroit AOD",
       x = "Founding lag (years from suburban crossing)",
       y = "Deferred maintenance ($M)", color = "Cohort")

aod_det %>%
  filter(!is.na(cohort)) %>%
  group_by(cohort) %>%
  summarise(
    n = n(),
    med_deferred_maint = median(deferred_maint_usd, na.rm = TRUE),
    med_net_income     = median(net_income_fy2425,  na.rm = TRUE),
    med_cash_savings   = median(cash_savings_usd,   na.rm = TRUE)
  ) %>%
  mutate(across(med_deferred_maint:med_cash_savings,
                ~ scales::dollar(., accuracy = 1))) %>%
  reactable(columns = list(
    cohort             = colDef(name = "Cohort"),
    n                  = colDef(name = "N"),
    med_deferred_maint = colDef(name = "Median deferred maint."),
    med_net_income     = colDef(name = "Median net income FY24/25"),
    med_cash_savings   = colDef(name = "Median cash savings")
  ), striped = TRUE, compact = TRUE)
```

```{r det-redevelopment-index, message=FALSE, warning=FALSE}
aod_index <- aod_det %>%
  filter(!is.na(deferred_maint_usd), !is.na(rev_fy2425),
         rev_fy2425 > 0, !is.na(pop_density)) %>%
  mutate(
    stress_score        = deferred_maint_usd / rev_fy2425,
    land_value_score    = pop_density,
    redevelopment_index = stress_score * land_value_score / 1e6
  )

top8 <- aod_index %>% slice_max(redevelopment_index, n = 8)

ggplot(aod_index, aes(x = land_value_score, y = stress_score,
                      size = deferred_maint_usd / 1e6, color = cohort)) +
  geom_point(alpha = 0.65) +
  geom_label_repel(data = top8, aes(label = short_name),
                   size = 2.5, max.overlaps = 20, seed = 42, box.padding = 0.4) +
  scale_color_manual(values = okabe_ito, na.value = "grey60") +
  scale_size_continuous(range = c(2, 8),
                        labels = scales::dollar_format(suffix = "M", prefix = "$")) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal() +
  labs(title = "Redevelopment pressure: financial stress × location density",
       subtitle = "Top 8 by redevelopment index labeled",
       x = "Population density (persons/km², Voronoi cell)",
       y = "Deferred maint. as % of annual revenue",
       size = "Deferred maint. ($M)", color = "Cohort")

aod_index %>%
  slice_max(redevelopment_index, n = 10) %>%
  select(short_name, cohort, deferred_maint_usd, rev_fy2425,
         stress_score, pop_density, redevelopment_index) %>%
  mutate(
    deferred_maint_usd  = scales::dollar(deferred_maint_usd / 1e3, accuracy = 1, suffix = "K"),
    rev_fy2425          = scales::dollar(rev_fy2425 / 1e3, accuracy = 1, suffix = "K"),
    stress_score        = scales::percent(stress_score, accuracy = 0.1),
    pop_density         = scales::comma(round(pop_density)),
    redevelopment_index = round(redevelopment_index, 3)
  ) %>%
  reactable(columns = list(
    short_name          = colDef(name = "Parish"),
    cohort              = colDef(name = "Cohort"),
    deferred_maint_usd  = colDef(name = "Deferred maint. ($K)"),
    rev_fy2425          = colDef(name = "Revenue FY24/25 ($K)"),
    stress_score        = colDef(name = "Stress ratio"),
    pop_density         = colDef(name = "Pop. density (persons/km²)"),
    redevelopment_index = colDef(name = "Redevelopment index")
  ), striped = TRUE, compact = TRUE)
```

```{r det-parking-cohort, message=FALSE, warning=FALSE}
parking_raw <- read_csv("overture_parking_results.csv", show_col_types = FALSE)

aod_parking <- aod_det %>%
  inner_join(parking_raw, by = "psuedo_id") %>%
  filter(overture_parking_area_m2 > 0) %>%
  mutate(
    rate_per_m2      = if_else(pop_density >= 1000, 1076, 538),
    implied_land_val = overture_parking_area_m2 * rate_per_m2
  )

ggplot(aod_parking, aes(x = overture_parking_area_m2,
                        y = reorder(short_name, overture_parking_area_m2),
                        color = cohort)) +
  geom_point(size = 3) +
  geom_segment(aes(x = 0, xend = overture_parking_area_m2,
                   yend = reorder(short_name, overture_parking_area_m2)),
               linewidth = 0.5) +
  scale_x_continuous(labels = scales::comma) +
  scale_color_manual(values = okabe_ito, na.value = "grey60") +
  theme_minimal() +
  labs(title = "Surface parking area, Detroit AOD parishes with Overture data",
       x = "Parking area (m²)", y = NULL, color = "Cohort")

cat(sprintf("Parishes with parking data: %d\n", nrow(aod_parking)))
cat(sprintf("Total implied surface parking land value: %s\n",
            scales::dollar(sum(aod_parking$implied_land_val, na.rm = TRUE), accuracy = 1000)))
cat(sprintf("Total deferred maintenance (same parishes): %s\n",
            scales::dollar(sum(aod_parking$deferred_maint_usd, na.rm = TRUE), accuracy = 1000)))
```

The [Redevelopment Analysis](Closure_Conversion_Recommendations.html) page translates this into a ranked national set of candidate sites, prioritizing parishes that simultaneously show high CLI, high surface parking area, and the territorial cohesion profile of a commuter-remnant parish.

---

## X. Conclusion

The canonical map is not simply outdated — it is actively maintained by a sacrificial logic that consumes individual parishes rather than reorganizing territorial boundaries. The Löschian framework tells us what efficient allocation of sacred services across space would look like, and measures a gap between that ideal and the canonical reality that is largest precisely in the metros where Catholicism was strongest. The Girardian framework tells us why that gap will not be closed through internal reform alone: the institution cannot acknowledge that the map is the problem without dissolving the differentiating order that gives the map its sacred authority. Each parish closure is not a rationalization — it is a sacrifice that preserves the fiction of the map by consuming one of its weakest nodes.

Cincinnati and Detroit together constitute a natural experiment in the *timing* of institutional sacrifice. Detroit has passed through the full sequence: misallocation → mimetic undifferentiation → sacrifice → void. Cincinnati is where Detroit was in approximately 2005: the same structural pattern, the same cohort gradient, the same null interaction between founding lag and rate of decline — but the restructuring process has not yet been triggered. The comparison isolates the AOD crisis as the amplifier that converts structural rigidity into institutional emergency, rather than the cause of the rigidity itself.

The structural finding — founding lag determines the level of sacramental activity, not the rate of decline — has a precise policy interpretation. The institution cannot accelerate its way out of this by making individual parishes more vital. The level effect is set by where the parish was founded relative to the demographic frontier; no amount of pastoral renewal changes the founding year. Reform requires either external demographic pressure sufficient to force territorial reorganization at scale, or a theological reframing that decouples canonical territory from sacred identity — that makes the map a convenience rather than a constitution.

---

## References

- Azzi, C., & Ehrenberg, R. (1975). Household Allocation of Time and Church Attendance. *Journal of Political Economy*, 83(1), 27–56.
- Berry, S. T. (1994). Estimating Discrete-Choice Models of Product Differentiation. *RAND Journal of Economics*, 25(2), 242–262.
- Conley, T. G. (1999). GMM Estimation with Cross Sectional Dependence. *Journal of Econometrics*, 92(1), 1–45.
- Finke, R., & Stark, R. (2005). *The Churching of America, 1776–2005.* Rutgers University Press.
- Girard, R. (1972). *La Violence et le Sacré.* Grasset. [Trans. *Violence and the Sacred*, 1977, Johns Hopkins UP.]
- Girard, R. (1982). *Le Bouc émissaire.* Grasset. [Trans. *The Scapegoat*, 1986, Johns Hopkins UP.]
- Hotelling, H. (1929). Stability in Competition. *The Economic Journal*, 39(153), 41–57.
- Huff, D. L. (1963). A Probabilistic Analysis of Shopping Center Trade Areas. *Land Economics*, 39(1), 81–90.
- Iannaccone, L. R. (1998). Introduction to the Economics of Religion. *Journal of Economic Literature*, 36(3), 1465–1495.
- Lösch, A. (1954). *The Economics of Location.* Yale University Press.
- McMillen, D. P. (2003). The Return of Centralization to Chicago. *Regional Science and Urban Economics*, 33(3), 287–304.
- National Council of Churches. (1956). *Churches and Church Membership in the United States, 1952.*
- Overture Maps Foundation. (2025). *Buildings and Infrastructure Layers: Technical Documentation.*
- Palaver, W. (2013). *René Girard's Mimetic Theory.* Michigan State University Press.
- Roman Catholic Archdiocese of Cincinnati. (2025). *Parish Vitality Report 2025.*
- Roman Catholic Archdiocese of Detroit. (2025–26). *Parish Restructuring Planning Workbooks.*
