---
title: "Shrinking Flock, Standing Stock? Catholic Decline vs. Real Estate in Cincinnati"
subtitle: "Exploratory notebook — not part of the published site, not a matched panel"
author: "Benedict Leonardi"
format:
  html:
    theme: journal
    toc: true
    number-sections: true
    code-fold: true
    self-contained: true
---

::: {.callout-warning}
## This is exploratory, not a finished result

This notebook pairs a long-run (1888-2021) diocese-level panel of Catholic
population/parishes against a **single present-day snapshot** of parcel-level
assessed real estate values for Catholic parishes in Hamilton County, Ohio
(the urban core of the Archdiocese of Cincinnati). Because the real estate
data has no historical time dimension, **this cannot be a matched
time-series/panel comparison** — it is a levels-vs-trend, mismatch-framing
exercise. Every number here also inherits the source data's own
documented limitations (see @sec-limitations). Treat conclusions as
suggestive, not a rigorous finding ready for the published papers in this
project.
:::

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(patchwork)
library(ggiraph)

okabe_ito <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
               "#0072B2", "#D55E00", "#CC79A7", "#000000")
```

## Data

### 1. Diocese panel (archdiocese-wide, 1888-2021)

```{r load-panel, message=FALSE}
panel <- read_csv("ocd_historical_panel/ocd_diocese_panel_merged.csv",
                   show_col_types = FALSE)

cin_raw <- panel %>%
  filter(canonical_diocese_name == "Cincinnati") %>%
  arrange(year)

# Plausibility filter, per the panel's own README/merge_log warnings:
# ~8.6% of total_catholic_population values nationally are implausible by
# magnitude alone (OCR digit-drop/repeat artifacts), and this failure mode
# is only weakly caught by the `quality_flags` column. Cincinnati's own
# known-good range across the 20th-21st century runs roughly 90k (1920,
# early growth phase) to ~550k (peak, late 1990s) -- so single-year values
# that collapse to a few thousand (e.g. 1986: 4,582; 2010: 4,682; 2019:
# 45,104) are dropped as OCR artifacts (a digit or two lost), not treated
# as real one-year population crashes. Same logic for total_parishes: the
# whole-panel run for Cincinnati sits in a tight 200s-250s band except one
# implausible single-year value (1999: 37, versus 232 in 2000 and 259 in
# 1970) which is dropped as a likely dropped-digit OCR error.
cin <- cin_raw %>%
  mutate(
    total_catholic_population = if_else(
      !is.na(total_catholic_population) &
        (total_catholic_population < 1.5e5 | total_catholic_population > 7e5),
      NA_real_, total_catholic_population
    ),
    total_parishes = if_else(
      !is.na(total_parishes) & total_parishes < 100,
      NA_real_, total_parishes
    )
  )

n_dropped_pop <- sum(!is.na(cin_raw$total_catholic_population)) - sum(!is.na(cin$total_catholic_population))
n_dropped_par <- sum(!is.na(cin_raw$total_parishes)) - sum(!is.na(cin$total_parishes))
```

`r n_dropped_pop` year(s) of `total_catholic_population` and `n_dropped_par`
year(s) of `total_parishes` were dropped as OCR-magnitude outliers (see code
above for the exact rule and rationale). This is a manual, judgment-call
filter on top of the panel's own `cleaned`/`merged` processing — appropriate
for a quick exploratory read, not a substitute for the systematic
`cleaning_log.md` methodology.

### 2. Parish real estate parcels (Hamilton County, present-day snapshot)

```{r load-parcels, message=FALSE}
parcels_raw <- read_csv("national_church_data/cincinnati_covington_church_parcels.csv",
                         show_col_types = FALSE)

# Restrict to the Archdiocese of Cincinnati AND Hamilton County, OH.
# Important scope note: the source file also carries Diocese of Covington
# (Kentucky side, a *different* diocese entirely) parishes, plus a couple of
# small Campbell/Kenton County KY rows -- those are dropped here. What's left
# is, in practice, essentially all of the "Cincinnati" rows in this file:
# the underlying dataset only captured Hamilton County for this diocese, not
# the full 19-county Archdiocese of Cincinnati footprint. That is a real
# geographic scope mismatch against the archdiocese-wide panel numbers above
# -- flagged again in the write-up below.
cin_parcels <- parcels_raw %>%
  filter(diocese_name == "Cincinnati", matched_county == "Hamilton_OH")

cin_parishes <- cin_parcels %>%
  filter(church_type == "Parish")

realestate_summary <- cin_parishes %>%
  summarise(
    n_parcels          = n(),
    n_distinct_parishes = n_distinct(church_name),
    land_value          = sum(parcel_taxable_land_value, na.rm = TRUE),
    improvement_value   = sum(parcel_taxable_improvement_value, na.rm = TRUE),
    total_value         = land_value + improvement_value,
    total_acres          = sum(parcel_area_acres, na.rm = TRUE)
  )
```

```{r rs-table}
realestate_summary %>%
  mutate(across(c(land_value, improvement_value, total_value), dollar)) %>%
  knitr::kable(col.names = c("Parcels", "Distinct parishes", "Land value",
                              "Improvement value", "Total value", "Acres"))
```

Hamilton County's `r realestate_summary$n_distinct_parishes` Catholic
parishes hold **`r dollar(realestate_summary$land_value)`** in aggregate
assessed *land* value alone, or **`r dollar(realestate_summary$total_value)`**
including improvements (buildings), across
`r round(realestate_summary$total_acres)` acres, per the county assessor
snapshot joined in `national_church_data/cincinnati_covington_church_parcels.csv`.
That's an average of roughly `r dollar(realestate_summary$total_value / realestate_summary$n_distinct_parishes)`
per parish in combined land+improvement value.

## The population/parish trend

```{r pop-trend}
pop_plot_df <- cin %>%
  filter(!is.na(total_catholic_population)) %>%
  mutate(tooltip = paste0("Archdiocese of Cincinnati, ", year,
                           "\nCatholic population: ", comma(total_catholic_population)))

pop_plot <- pop_plot_df %>%
  ggplot(aes(year, total_catholic_population)) +
  geom_line_interactive(color = okabe_ito[5], linewidth = 0.8) +
  geom_point_interactive(aes(tooltip = tooltip, data_id = year),
                          color = okabe_ito[5], size = 1.8) +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  labs(
    title = "Archdiocese of Cincinnati: Catholic population, 1888-2021",
    subtitle = "OCD-derived panel, outlier years dropped (see notes above)",
    x = NULL, y = "Reported Catholic population"
  ) +
  theme_minimal(base_size = 12)

parish_plot_df <- cin %>%
  filter(!is.na(total_parishes)) %>%
  mutate(tooltip = paste0("Archdiocese of Cincinnati, ", year,
                           "\nTotal parishes: ", comma(total_parishes)))

parish_plot <- parish_plot_df %>%
  ggplot(aes(year, total_parishes)) +
  geom_line_interactive(color = okabe_ito[6], linewidth = 0.8) +
  geom_point_interactive(aes(tooltip = tooltip, data_id = year),
                          color = okabe_ito[6], size = 1.8) +
  labs(
    title = "Archdiocese of Cincinnati: total parishes, 1888-2021",
    subtitle = "Sparser field (0.5% national parse rate); points shown only where reported",
    x = NULL, y = "Reported parish count"
  ) +
  theme_minimal(base_size = 12)

girafe(ggobj = pop_plot / parish_plot, width_svg = 8, height_svg = 8) %>%
  girafe_options(opts_sizing(rescale = TRUE), opts_hover(css = "stroke-width:2;"))
```

```{r pop-stats}
peak_row   <- cin %>% filter(!is.na(total_catholic_population)) %>% slice_max(total_catholic_population, n = 1, with_ties = FALSE)
latest_row <- cin %>% filter(!is.na(total_catholic_population)) %>% slice_max(year, n = 1)
early_row  <- cin %>% filter(!is.na(total_catholic_population), year <= 1965) %>% slice_max(year, n = 1)

pct_decline_from_peak  <- 1 - latest_row$total_catholic_population / peak_row$total_catholic_population
pct_decline_since_1960s <- 1 - latest_row$total_catholic_population / early_row$total_catholic_population
```

Reading the filtered series: reported Catholic population climbs from
~91,000 (1920) to a plausible-series peak of
`r comma(peak_row$total_catholic_population)` in `r peak_row$year`, then
declines to `r comma(latest_row$total_catholic_population)` by
`r latest_row$year` — a **`r percent(pct_decline_from_peak, accuracy = 1)`
decline from peak**, or **`r percent(pct_decline_since_1960s, accuracy = 1)`**
measured from the `r early_row$year` reading
(`r comma(early_row$total_catholic_population)`) to `r latest_row$year`.
(The 2021 raw value, ~305k, was left in the filtered series since it passes
the magnitude filter, but note it carries several *other* field quality
flags in that row — treat the very last data point with more caution than
the rest of the line.)

Parish counts, where reported, run from 219 (1920) up to 259 (1970) and back
down to 218 by 2010 — a much shallower decline (roughly 16% off the 1970
peak) than the population decline over a comparable window. That gap itself
— population falling faster than parish count — is the setup for the real
estate question: if parishes aren't closing in proportion to membership
loss, the underlying real estate isn't being shed either.

## Putting the two together

```{r bar-chart}
bar_df <- tibble(
  metric = c("Catholic population\n(1998 peak → 2018)", "Parishes\n(1970 peak → 2010)"),
  pct_change = c(-pct_decline_from_peak, -(259 - 218) / 259)
) %>%
  mutate(tooltip = paste0(str_replace_all(metric, "\n", " "),
                           ": ", percent(pct_change, accuracy = 1), " change from peak"))

p_bar <- ggplot(bar_df, aes(metric, pct_change)) +
  geom_col_interactive(aes(tooltip = tooltip, data_id = metric), fill = okabe_ito[6], width = 0.5) +
  geom_hline(yintercept = 0) +
  geom_text(aes(label = percent(pct_change, accuracy = 1)), vjust = 1.4, color = "white", fontface = "bold") +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Demand has fallen faster than supply of parishes",
    subtitle = "Archdiocese-wide, from each series' own peak to its latest reliable reading",
    x = NULL, y = "% change from peak"
  ) +
  theme_minimal(base_size = 12)

girafe(ggobj = p_bar, width_svg = 8, height_svg = 5) %>%
  girafe_options(opts_sizing(rescale = TRUE), opts_hover(css = "stroke-width:2;"))
```

The headline pairing, stated carefully:

- **Archdiocese-wide** Catholic population fell about
  `r percent(pct_decline_from_peak, accuracy = 1)` from its reported peak
  (`r peak_row$year`) to the most recent plausible reading
  (`r latest_row$year`).
- **Archdiocese-wide** parish count fell only ~16% from its 1970 peak (259)
  to 2010 (218) — a much shallower decline.
- **Hamilton-County-only**, present day: `r realestate_summary$n_distinct_parishes`
  parishes still hold `r dollar(realestate_summary$total_value)` in combined
  assessed land+improvement value (`r dollar(realestate_summary$land_value)`
  land alone).

There is no way, with this data, to say whether that `r dollar(realestate_summary$total_value)`
figure is *up or down* in real terms from 1970 or 1998 — there is no
historical assessor series in hand for this project (see @sec-limitations).
But the **structural** picture is consistent with an "excess real estate"
story: the archdiocese has shed membership much faster than it has shed
parishes, so whatever the trajectory of land values, the *number of
separately-held parcels* generating carrying costs (insurance, utilities,
deferred maintenance, property tax on non-exempt or mixed-use parcels) has
not shrunk in step with demand. This is the same mismatch that
`Economics_of_Church_Consolidation.qmd` argues from a present-day snapshot;
this notebook adds the demand-side historical depth but does not (and
currently cannot) add a matching historical supply-side value series.

A secondary observation worth flagging: the Hamilton County parcel file
includes a non-trivial amount of parish-held real estate that is not the
church building itself — `r sum(cin_parishes$parcel_land_use_category == "Vacant Land", na.rm = TRUE)`
parcels are classified "Vacant Land," and dozens more are single-family,
multi-family, or commercial/retail parcels (rectories, former convents/
schools, adjacent lots) rather than "Church/Religious" use. That's
consistent with parishes sitting on more land than their current
sacramental footprint requires — exactly the kind of latent real estate
this project's consolidation thesis is about — but is not itself evidence
about value *trends*.

## Limitations {#sec-limitations}

1. **No historical real estate series.** The parcel file is a single
   present-day Hamilton County assessor snapshot. There is no equivalent
   1970s/1990s assessed-value series for these same parcels anywhere in this
   project, and none was fetched for this notebook per instructions (existing
   Census/NHGIS extracts already in this project —
   `nhgis0001_csv`/`nhgis0002_csv` — turned out to contain only 1950/1970
   *population density* tables, not housing/property value variables, so they
   could not be repurposed for a real-estate trend check without a fresh
   data pull). This means the central question the task poses — "is real
   estate value declining proportionally with membership" — **cannot
   actually be answered with a real trend comparison here.** What's shown
   instead is a *levels* comparison: today's real estate stock against the
   long-run membership trend.
2. **Geographic scope mismatch.** The diocese panel figures (population,
   parishes) are archdiocese-wide, covering the full ~19-county Archdiocese
   of Cincinnati footprint in southwest Ohio. The parcel file, despite being
   labeled "Cincinnati," only captured Hamilton County (the urban core) for
   this diocese. Hamilton County is where the oldest, most population-loss-
   exposed urban parishes are concentrated, so if anything this likely
   *overstates* the mismatch between demand decline and real estate
   persistence relative to the full archdiocese (suburban/exurban counties
   in the same archdiocese have plausibly seen parish growth, not decline,
   over the same period — consistent with this project's own
   `Suburbanization_Analysis.qmd` finding a suburban shift in parish
   footprint).
3. **`total_catholic_population` and `total_parishes` are both
   OCR-extraction-derived and imperfect even after cleaning/merging.** Per
   `ocd_historical_panel/README.md`, roughly 8.6% of national
   `total_catholic_population` values are magnitude-implausible even before
   any manual eyeballing, and this notebook's own outlier filter (a simple
   magnitude bound, not the full documented `cleaning_log.md` methodology)
   is a coarser, faster pass than the project's systematic cleaning process.
   `total_parishes` has only a 0.5% national parse rate, so the parish-count
   line above is sparse and each point should be read as "reported in a year
   OCD editions happened to include a parseable statistics block," not a
   dense annual series.
4. **"Assessed land value" is a tax-assessment construct, not a market
   appraisal.** All rows in the Hamilton County parcel file show
   `parcel_is_fully_exempt == FALSE`, which is unexpected for church
   property (churches are typically property-tax exempt) — this may mean
   the assessor's underlying land/improvement values used here are a
   pre-exemption or reference valuation rather than the parishes' true
   current market value; this was not independently verified against
   Hamilton County's own exemption rules for this notebook.
5. **Not a matched entity list.** The `r peak_row$year` vs. `r latest_row$year`
   population comparison and the present-day parish real estate total are
   not drawn from an identical, fixed list of parishes — parishes that
   closed/merged/opened between those dates are not separately tracked here.
