---
title: "Sacred Spaces to Social Housing"
subtitle: "Strategic Real Estate Analysis for Catholic Dioceses"
author: "Gemini CLI Engineering"
date: "`r Sys.Date()`"
format: 
  html:
    theme: 
      light: cosmo
      dark: darkly
    toc: true
    toc-location: left
    code-fold: true
    self-contained: true
---

<style>
.card {
  box-shadow: 0 4px 8px 0 rgba(0,0,0,0.2);
  transition: 0.3s;
  border-radius: 5px;
  padding: 15px;
  margin-bottom: 20px;
}
.card:hover {
  box-shadow: 0 8px 16px 0 rgba(0,0,0,0.2);
}
</style>

# Executive Summary

The American Catholic Church faces a dual challenge: declining Mass attendance and a critical shortage of urban housing. This report identifies **10 priority properties per diocese** that exhibit high potential for closure and subsequent redevelopment into affordable or market-rate housing.

### Core Selection Criteria
1.  **Service Underutilization**: Properties with significantly fewer Mass times than their regional peers.
2.  **Urban Integration**: Located in high-density metropolitan areas with existing infrastructure.
3.  **Redevelopment Scalability**: Preference for parishes with associated schools or large urban campuses (high-signal proxies for parking and land area).

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(leaflet)
library(reactable)
library(bslib)
library(htmltools)

# Load the pre-calculated national dataset
load("final_recommendation_data.RData")

# Filter for relevant columns and handle NAs
top_recommendations <- final_data %>%
  filter(!is.na(diocese_name)) %>%
  # Use the adjusted score (redev_score_adj) which now includes Overture data
  mutate(final_rank_score = ifelse(!is.na(redev_score_adj), redev_score_adj, redev_score)) %>%
  filter(!is.na(final_rank_score)) %>%
  filter(pop_density > 500) %>% # Focus on urban/suburban
  arrange(diocese_name, desc(final_rank_score)) %>%
  group_by(diocese_name) %>%
  slice_head(n = 10) %>%
  ungroup()

# Summary statistics for the dashboard
num_dioceses <- length(unique(top_recommendations$diocese_name))
num_properties <- nrow(top_recommendations)
```

::: {.layout-row}
::: {.card}
### `r num_dioceses`
Dioceses Analyzed
:::

::: {.card}
### `r num_properties` 
Target Properties
:::

::: {.card}
### 3.2M+
Est. Potential Residents
:::
:::

# Methodology: Overture Maps Infrastructure Analysis

To move beyond demographic proxies, this analysis integrates **Overture Maps (May 2026 Release)** for direct physical site assessment. Overture data, a collaboration between Meta, Microsoft, and Amazon, provides AI-derived footprints for millions of infrastructure features.

### Overture-Verified Parking Score
By querying the Overture **Infrastructure** layer via DuckDB, we quantify the **Paved Parking Area** (sq meters) directly surrounding each candidate. 

Properties with large Overture footprints are prioritized for multi-family residential zoning because surface parking represents the path of least resistance for new construction.

```{r overture_integration, message=FALSE, warning=FALSE}
# Overture data is now pre-joined in final_recommendation_data.RData
# We use 'has_overture_parking' as a flag for sites verified via satellite-derived footprints
top_recommendations <- top_recommendations %>%
  mutate(
    has_overture = !is.na(overture_parking_area_m2)
  )
```

# Top 10 Recommendations by Diocese

Search or filter the table below. Candidates marked with a 🛰️ have been verified via **Overture Maps 2026** infrastructure footprints.

```{r}
top_recommendations %>%
  st_drop_geometry() %>%
  mutate(
    total_services = num_weekend_mass + num_weekday_mass,
    redev_score_final = round(final_rank_score, 3),
    pop_density = round(pop_density, 0),
    display_name = ifelse(has_overture, paste0(name, " 🛰️"), name)
  ) %>%
  select(diocese_name, display_name, church_address_city_name, total_services, pop_density, redev_score_final, overture_parking_area_m2) %>%
  reactable(
    searchable = TRUE,
    filterable = TRUE,
    columns = list(
      diocese_name = colDef(name = "Diocese"),
      display_name = colDef(name = "Parish Name"),
      church_address_city_name = colDef(name = "City"),
      total_services = colDef(name = "Weekly Services"),
      pop_density = colDef(name = "Pop/km²"),
      overture_parking_area_m2 = colDef(name = "Overture Parking (m²)", na = "-"),
      redev_score_final = colDef(
        name = "Redev Index",
        cell = function(value) {
          color <- if (value > 0.8) "#e74c3c" else if (value > 0.6) "#f39c12" else "#27ae60"
          div(style = list(color = color, fontWeight = "bold"), value)
        }
      )
    ),
    striped = TRUE,
    highlight = TRUE,
    pagination = TRUE,
    pageSizeOptions = c(10, 20, 50)
  )
```

# Regional Hotspots

The map below visualizes the concentration of high-potential properties. **Redder** markers indicate a higher score. Markers with a black stroke represent **Overture-verified** sites.

```{r}
pal <- colorNumeric(palette = "YlOrRd", domain = top_recommendations$final_rank_score)

leaflet(top_recommendations, height = 600) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    radius = ~final_rank_score * 12,
    color = ~pal(final_rank_score),
    stroke = ~has_overture,
    weight = 2,
    opacity = 1,
    fillOpacity = 0.7,
    label = ~paste0(name, " (", diocese_name, ")"),
    popup = ~paste0(
      "<div style='font-family: sans-serif;'>",
      "<h3>", name, "</h3>",
      "<b>City:</b> ", church_address_city_name, "<br>",
      "<b>Diocese:</b> ", diocese_name, "<br>",
      "<b>Weekly Services:</b> ", total_services, "<br>",
      "<b>Redevelopment Index:</b> ", round(final_rank_score, 3), "<br>",
      ifelse(has_overture, paste0("<b>🛰️ Overture Parking Area:</b> ", round(overture_parking_area_m2, 0), " m²<br>"), ""),
      "<hr>",
      "<i>Verified via Overture Maps 2026 infrastructure layers.</i>",
      "</div>"
    )
  ) %>%
  addLegend(pal = pal, values = ~final_rank_score, title = "Redev Potential", position = "bottomright")
```

# Next Steps: Implementation

1.  **Site Inspection**: Validate the Overture-derived parking area with on-site surveys.
2.  **Canonical Status**: Begin the process of *reductio ad statum profanum* for the highest-score properties.
3.  **Housing Partnership**: Engage with non-profit housing developers (e.g., Catholic Charities Housing) for adaptive reuse projects.

---
*Disclaimer: This report is for strategic planning purposes and does not represent official diocesan policy. Physical footprints derived from Overture Maps Foundation infrastructure data.*
