---
title: "Redevelopment Analysis: Priority Sites by Diocese"
subtitle: "Identifying Underutilized Urban Church Land for Potential Conversion"
author: "Benedict Leonardi"
date: "`r Sys.Date()`"
---

## Overview

The [main analysis](Canonical_Rigidity_Research_Paper.html) identifies the structural conditions under which canonical rigidity generates asset misallocation: parishes in high-density urban areas holding large surface parking footprints while offering fewer services than their surrounding population would predict. This page translates that framework into a ranked set of candidate sites.

The **Redevelopment Index** is a composite score across three dimensions:

$$\text{Redev Score}_i = f\!\left(\text{PopDensity}_i,\; \text{ParkingArea}_i,\; \frac{1}{\text{MassFrequency}_i}\right)$$

Higher scores indicate greater misalignment between current institutional use and the potential value of the underlying land. Parishes are filtered to urban and suburban density tiers (>500 persons/km²) to focus on areas where the opportunity cost of surface parking is meaningful.

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(leaflet)
library(reactable)
library(ggiraph)

load("final_recommendation_data.RData")

top_recommendations <- final_data %>%
  filter(!is.na(diocese_name)) %>%
  mutate(final_rank_score = ifelse(!is.na(redev_score_adj), redev_score_adj, redev_score)) %>%
  filter(!is.na(final_rank_score), pop_density > 500) %>%
  arrange(diocese_name, desc(final_rank_score)) %>%
  group_by(diocese_name) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  mutate(
    total_services  = num_weekend_mass + num_weekday_mass,
    has_overture    = !is.na(overture_parking_area_m2),
    display_name    = ifelse(has_overture, paste0(name, " *"), name)
  )

num_dioceses  <- length(unique(top_recommendations$diocese_name))
num_properties <- nrow(top_recommendations)
```

**`r num_dioceses` dioceses · `r num_properties` candidate properties** meeting the urban/suburban density threshold. Sites marked with * have parking area verified via Overture Maps 2026 infrastructure footprints.

---

## Priority Sites by Diocese

```{r table}
top_recommendations %>%
  st_drop_geometry() %>%
  mutate(
    redev_score_final = round(final_rank_score, 3),
    pop_density       = round(pop_density, 0)
  ) %>%
  select(diocese_name, display_name, church_address_city_name,
         total_services, pop_density, redev_score_final, overture_parking_area_m2) %>%
  reactable(
    searchable  = TRUE,
    filterable  = TRUE,
    striped     = TRUE,
    highlight   = TRUE,
    pagination  = TRUE,
    pageSizeOptions = c(10, 20, 50),
    columns = list(
      diocese_name              = colDef(name = "Diocese"),
      display_name              = colDef(name = "Parish"),
      church_address_city_name  = colDef(name = "City"),
      total_services            = colDef(name = "Weekly Services"),
      pop_density               = colDef(name = "Pop/km²"),
      overture_parking_area_m2  = colDef(name = "Parking Area (m²)", na = "—"),
      redev_score_final         = colDef(
        name = "Redev Index",
        cell = function(value) {
          color <- if (value > 0.8) "#c0392b" else if (value > 0.6) "#e67e22" else "#27ae60"
          htmltools::div(style = list(color = color, fontWeight = "bold"), value)
        }
      )
    )
  )
```

---

## Geographic Distribution

Marker size scales to the Redevelopment Index score; sites with Overture-verified parking footprints are shown with a dark border.

```{r map}
pal <- colorNumeric(palette = "YlOrRd", domain = top_recommendations$final_rank_score)

leaflet(top_recommendations, height = 580) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    radius     = ~final_rank_score * 12,
    fillColor  = ~pal(final_rank_score),
    color      = ~ifelse(has_overture, "#2c3e50", "#888888"),
    weight     = ~ifelse(has_overture, 2, 0.5),
    fillOpacity = 0.75,
    label = ~paste0(name, " (", diocese_name, ")"),
    popup = ~paste0(
      "<b>", name, "</b><br>",
      diocese_name, " · ", church_address_city_name, "<br>",
      "Weekly services: ", total_services, "<br>",
      "Pop density: ", round(pop_density), " /km²<br>",
      "Redev index: ", round(final_rank_score, 3),
      ifelse(has_overture,
             paste0("<br>Parking: ", round(overture_parking_area_m2), " m² (Overture verified)"),
             "")
    )
  ) %>%
  addLegend(pal = pal, values = ~final_rank_score,
            title = "Redev Index", position = "bottomright")
```

---

## Detroit AOD: Ranked Redevelopment Candidates

The Archdiocese of Detroit's 2025–26 restructuring process (*Acts of the Disciples*) produced parish workbooks for all 209 active parishes containing financial data not available in the national dataset — deferred maintenance costs, fiscal year 2024–25 annual revenue, and cash savings balances. Combining these figures with Voronoi-based population density estimates allows a more precise redevelopment ranking for this diocese: rather than relying on a proxy composite, we can compute a direct financial-stress metric and weight it by the population-density opportunity cost of the underlying land.

```{r det-redev-table}
library(scales)

aod_raw <- read_csv("aod_voronoi_data.csv", show_col_types = FALSE)

det_scored <- aod_raw %>%
  mutate(
    stress_score       = ifelse(!is.na(deferred_maint_usd) & !is.na(rev_fy2425) & rev_fy2425 > 0,
                                deferred_maint_usd / rev_fy2425, NA_real_),
    redevelopment_index = stress_score * pop_density / 1e6
  ) %>%
  filter(!is.na(stress_score), !is.na(redevelopment_index), pop_density > 500) %>%
  arrange(desc(redevelopment_index))

q75 <- quantile(det_scored$redevelopment_index, 0.75, na.rm = TRUE)
q50 <- quantile(det_scored$redevelopment_index, 0.50, na.rm = TRUE)

det_scored %>%
  slice_head(n = 25) %>%
  select(short_name, city, cohort,
         deferred_maint_usd, rev_fy2425,
         stress_score, pop_density, redevelopment_index) %>%
  reactable(
    searchable  = TRUE,
    filterable  = TRUE,
    striped     = TRUE,
    highlight   = TRUE,
    columns = list(
      short_name          = colDef(name = "Parish"),
      city                = colDef(name = "City"),
      cohort              = colDef(name = "Cohort"),
      deferred_maint_usd  = colDef(
        name = "Deferred Maint.",
        cell = function(v) scales::dollar(v, scale = 1e-3, suffix = "K")
      ),
      rev_fy2425          = colDef(
        name = "Annual Revenue",
        cell = function(v) scales::dollar(v, scale = 1e-3, suffix = "K")
      ),
      stress_score        = colDef(
        name = "Stress Score",
        cell = function(v) round(v, 2)
      ),
      pop_density         = colDef(
        name = "Pop/km²",
        cell = function(v) round(v, 0)
      ),
      redevelopment_index = colDef(
        name = "Redev Index",
        cell = function(value) {
          color <- if (value > q75) "#c0392b" else if (value > q50) "#e67e22" else "#27ae60"
          htmltools::div(style = list(color = color, fontWeight = "bold"), round(value, 3))
        }
      )
    )
  )
```

```{r det-redev-map}
det_map <- det_scored %>%
  filter(!is.na(latitude), !is.na(longitude))

pal_det <- colorNumeric(palette = "YlOrRd", domain = det_map$redevelopment_index)

leaflet(det_map, height = 520) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    lng         = ~longitude,
    lat         = ~latitude,
    radius      = ~redevelopment_index * 12,
    fillColor   = ~pal_det(redevelopment_index),
    color       = "#555555",
    weight      = 0.8,
    fillOpacity = 0.75,
    popup = ~paste0(
      "<b>", short_name, "</b><br>",
      "Cohort: ", cohort, "<br>",
      "Deferred maint.: $", round(deferred_maint_usd / 1e3), "K<br>",
      "Stress score: ", round(stress_score, 2), "<br>",
      "Pop density: ", round(pop_density), " /km²<br>",
      "Redev index: ", round(redevelopment_index, 3)
    )
  ) %>%
  addLegend(pal = pal_det, values = ~redevelopment_index,
            title = "Redev Index", position = "bottomright")
```

```{r det-parking-shadow}
parking <- read_csv("overture_parking_results.csv", show_col_types = FALSE)

det_parking <- aod_raw %>%
  inner_join(parking, by = "psuedo_id") %>%
  filter(!is.na(overture_parking_area_m2), overture_parking_area_m2 > 0) %>%
  mutate(
    shadow_rate     = ifelse(pop_density >= 1000, 1076, 538),
    shadow_rent_usd = overture_parking_area_m2 * shadow_rate
  )

total_land_value  <- sum(det_parking$shadow_rent_usd, na.rm = TRUE)
n_parishes        <- nrow(det_parking)
med_deferred      <- median(det_parking$deferred_maint_usd, na.rm = TRUE)

cat(sprintf(
  "Parishes with parking data: %d\nTotal implied parking land value: $%s\nMedian deferred maintenance: $%s\n",
  n_parishes,
  scales::dollar(total_land_value),
  scales::dollar(med_deferred)
))

det_parking <- det_parking %>%
  mutate(tooltip = paste0(short_name, "\nCohort: ", cohort,
                           "\nImplied parking land value: $", round(shadow_rent_usd / 1e6, 2), "M"))

p_shadow_rent <- ggplot(det_parking,
       aes(x = reorder(short_name, shadow_rent_usd),
           y = shadow_rent_usd / 1e6,
           color = cohort)) +
  geom_segment(aes(xend = short_name, yend = 0), linewidth = 0.6) +
  geom_point_interactive(aes(tooltip = tooltip, data_id = short_name), size = 3) +
  geom_hline(yintercept = med_deferred / 1e6,
             linetype = "dashed", color = "#c0392b", linewidth = 0.8) +
  annotate("text", x = 1, y = med_deferred / 1e6,
           label = "Median deferred maintenance",
           vjust = -0.5, hjust = 0, size = 3, color = "#c0392b") +
  coord_flip() +
  scale_y_continuous(labels = scales::dollar_format(suffix = "M")) +
  labs(
    x     = NULL,
    y     = "Implied Parking Land Value ($M)",
    color = "Cohort",
    title = "Shadow Rent on AOD Parking Footprints vs. Deferred Maintenance"
  ) +
  theme_minimal(base_size = 11)

girafe(ggobj = p_shadow_rent, width_svg = 8, height_svg = 6) %>%
  girafe_options(opts_sizing(rescale = TRUE), opts_hover(css = "stroke-width:2;"))
```

---

## On the Policy Logic

The redevelopment argument is not that churches should close — it is that the specific sites flagged here represent a category where the mismatch between institutional use and urban land economics is particularly acute. A parish offering 6 weekend Masses in a neighborhood with 4,000 residents per km² and 5,000 m² of surface parking is, from a land-use standpoint, an outlier. What exactly should happen to that site involves canonical, historic preservation, and community considerations well outside the scope of this analysis.

The more tractable proposal is the *ground-lease model*: dioceses retain ownership of sacred structures while leasing surface parking to developers for multi-family residential construction. This preserves the parish's sacramental function, removes no seats from the nave, and converts dead land into a stable endowment-style revenue stream — a structure that Catholic Charities Housing and similar organizations have piloted in several dioceses.

---

*Infrastructure footprints from Overture Maps Foundation (May 2026 release). Revenue and population data from ACS 2022 5-Year Estimates.*
