---
title: "The Spatial Economics of Sacred Space: Market Consolidation and Real Estate Optimization in the US Catholic Church"
author: 
  - name: "Benedict Leonardi"
    affiliations: "Independent Researcher"
date: "`r Sys.Date()`"
abstract: |
  This paper examines the spatial distribution, market area efficiency, and real estate footprint of the Catholic Church in the United States. Operating as a territorial monopoly regulated by canon law, the Church’s legacy infrastructure frequently misaligns with contemporary demographic shifts and declining religious adherence. Utilizing a novel dataset of over 23,000 geolocated parishes, integrated with service frequency (Mass times), US Census demographic data, and AI-derived Overture Maps infrastructure footprints, we model the "Spiritual Landscape" through the lens of spatial economics. We first employ Voronoi tessellations to identify geographic market catchment areas, revealing significant spatial "leakage" where parishioner proximity conflicts with rigid ecclesiastical boundaries. Second, we develop a Real Estate Redevelopment Index to identify heavily underutilized urban assets. By targeting parishes with low service density, high local population density, and extensive surface parking, we identify priority properties across US dioceses that are optimal for multi-family residential conversion. We estimate that strategically liquidating or repurposing these specific underperforming assets could alleviate local housing shortages while simultaneously stabilizing diocesan balance sheets.
format: 
  html:
    theme: journal
    toc: true
    number-sections: true
    code-fold: true
    self-contained: true
---

## Introduction

The Roman Catholic Church in the United States represents one of the largest private holders of real estate, operating a vast network of parishes, schools, and administrative buildings. Economically, the Church functions as a localized spatial monopoly. Under Canon Law, geographic territories are strictly delineated into parishes, aggregated into dioceses. Parishioners are legally bound to their geographic parish. However, the modern mobility of the American public, combined with changing demographics and declining weekly religious adherence, has fractured this rigid model. 

Today, the institutional Church faces two converging macroeconomic realities:
1. **Demand-Side Contraction:** A long-term secular decline in weekly Mass attendance and clerical vocations, resulting in excess capacity (underutilized physical space).
2. **Urban Real Estate Scarcity:** A severe housing shortage in major metropolitan areas, driving up land values and increasing the opportunity cost of holding non-revenue-generating surface parking and underutilized ecclesiastical buildings.

This paper applies spatial economics and urban land-use theory to analyze this mismatch. We seek to answer two primary questions: How closely do legal ecclesiastical boundaries align with actual geographic market areas (spatial accessibility)? And secondly, how can dioceses mathematically optimize their real property portfolios by identifying underutilized assets for redevelopment?

## Literature Review

This analysis bridges three sub-fields of economics:

**1. The Economics of Religion:** Foundational work by Azzi and Ehrenberg (1975) framed religious participation as an intertemporal allocation of time and resources. Iannaccone (1998) expanded this to view denominations as firms competing in a religious marketplace. While Protestant denominations frequently behave as firms in a monopolistically competitive market, the Catholic Church operates more akin to a franchisor assigning strict regional monopolies (Dioceses/Parishes).

**2. Spatial Competition and Market Areas:** Following Hotelling (1929) and Losch (1954), spatial economics dictates that consumers will travel to the nearest provider to minimize transportation costs. When administrative boundaries (e.g., diocesan borders) force consumers to bypass closer options, it creates spatial friction or "leakage." 

**3. Urban Land Use and the Opportunity Cost of Space:** The Alonso-Muth-Mills monocentric city model dictates that land closer to high-density urban centers demands higher rents and denser capital development. Surface parking lots attached to urban parishes represent a severe misallocation of land in high-density zones (Shoup, 2005).

## Data

Our analysis relies on a novel synthesis of four distinct datasets:
*   **Institutional Data:** Geocoded locations of approximately 23,000 Catholic institutions scraped from national directories (`national_masstimes_scrape.csv`).
*   **Service Demand Proxies:** Ritual frequency data including the number of weekend and weekday Masses per parish (`masstimes_national_data.csv`), serving as a proxy for institutional vitality and localized demand.
*   **Demographic Data:** US Census American Community Survey (ACS) tract-level data, capturing population density, median age, and socioeconomic variables.
*   **Infrastructure Data (Overture Maps):** May 2026 release data from Overture Maps (a Meta, Microsoft, and Amazon collaboration) queried via DuckDB. We extract the physical footprints of paved parking areas (`overture_parking_area_m2`) surrounding each church.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(leaflet)

# Load core institutional dataset
church_data <- read_csv("national_masstimes_scrape.csv", show_col_types = FALSE) %>%
  filter(church_address_country_territory_name == "United States", !is.na(latitude))

# Load pre-processed real estate redevelopment data
load("final_recommendation_data.RData")
```

## Methodology

### 1. Geometric Parochiality and Spatial Leakage

To estimate the "natural" market area of a parish irrespective of its legal boundaries, we compute Voronoi tessellations across the United States. A Voronoi diagram partitions a plane into regions based on the distance to a specific set of points (churches). The area within a parish's Voronoi polygon represents the territory where that specific parish is geographically the closest option.

We define "Spatial Leakage" as regions where a parishioner's Voronoi polygon belongs to a church in a different diocese than their physical residence. 

### 2. The Redevelopment Index

To identify optimal targets for closure, consolidation, or conversion, we construct a **Real Estate Redevelopment Index**. The index is a composite score maximizing the following attributes:
*   **Underutilization (Service Scarcity):** Total weekly masses (weekend + weekday). Lower frequencies increase the score.
*   **Zoning Potential (Overture Parking):** The square meterage of surface parking. Large surface lots in urban cores represent the "path of least resistance" for multi-family residential zoning, as they do not require demolition of historic church structures.
*   **Market Demand (Population Density):** Tract-level population density. Higher density indicates stronger market demand for converted housing.

$$ \text{Redevelopment Score}_i = f(\text{PopDensity}_i, \text{ParkingArea}_i, \frac{1}{\text{MassFrequency}_i}) $$

## Results & Discussion

### The Spatial Friction of Canonical Borders

The spatial analysis reveals profound inefficiencies in the Church's territorial model. Canonical borders frequently trace natural geographic features (like rivers) or state lines, ignoring modern transportation networks (highways, bridges) that dictate true consumer travel costs.

**Case Study: Cincinnati (OH) vs. Covington (KY)**
Separated by the Ohio River, the Archdiocese of Cincinnati and the Diocese of Covington represent distinct legal markets. However, Voronoi analysis shows substantial overlap. Many residents in Northern Kentucky are spatially closer to downtown Cincinnati parishes than to their own diocesan alternatives. This geometric reality encourages "parish shopping" and undermines the monopolistic pricing power (tithing expectations) of the local diocese.

```{r cin_cov_analysis, message=FALSE, warning=FALSE}
cin_cov <- church_data %>%
  filter(diocese_name %in% c("Cincinnati", "Covington")) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32616)

# Generate Voronoi Polygons
voronoi_cin_cov <- st_voronoi(st_union(cin_cov)) %>%
  st_collection_extract() %>%
  st_sf() %>%
  st_join(cin_cov) %>%
  st_transform(4326)

pal <- colorFactor(palette = c("#0072B2", "#D55E00"), domain = cin_cov$diocese_name)

leaflet(voronoi_cin_cov) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(fillColor = ~pal(diocese_name), fillOpacity = 0.4, color = "white", weight = 1) %>%
  addLegend(pal = pal, values = ~diocese_name, title = "Diocese Spatial Area")
```

### Strategic Asset Optimization (The Top 10 Portfolios)

Applying the Redevelopment Index across the `final_recommendation_data`, we filter for high-density urban areas (> 500 pop/km²) to identify the top 10 redevelopment targets per diocese. 

Properties scoring highly on this index operate fewer than the regional average of weekend services but sit on extensive, Overture-verified paved footprints in high-rent districts. Consolidating these parishes and leasing the land to developers for multi-family housing presents a massive economic opportunity.

```{r redev_table, message=FALSE, warning=FALSE}
library(reactable)
library(htmltools)

top_targets <- final_data %>%
  filter(!is.na(diocese_name), pop_density > 500, !is.na(redev_score_adj)) %>%
  arrange(diocese_name, desc(redev_score_adj)) %>%
  group_by(diocese_name) %>%
  slice_head(n = 3) %>% # Showing top 3 per diocese for brevity in paper
  ungroup() %>%
  mutate(
    total_services = num_weekend_mass + num_weekday_mass,
    display_name = ifelse(!is.na(overture_parking_area_m2), paste0(name, " 🛰️"), name)
  ) %>%
  select(diocese_name, display_name, church_address_city_name, total_services, pop_density, overture_parking_area_m2)

reactable(
  top_targets,
  columns = list(
    diocese_name = colDef(name = "Diocese"),
    display_name = colDef(name = "Parish Name"),
    church_address_city_name = colDef(name = "City"),
    total_services = colDef(name = "Weekly Masses"),
    pop_density = colDef(name = "Pop/km²"),
    overture_parking_area_m2 = colDef(name = "Parking Space (m²)", na = "N/A")
  ),
  striped = TRUE, highlight = TRUE
)
```
*(Note: The 🛰️ icon indicates surface parking was verified via 2026 AI-derived Overture Maps satellite footprints).*

## Historical Diocesan Trajectories: Excess Parish Supply Over Time

The Redevelopment Index above is a purely cross-sectional, parish-level snapshot: it says nothing about *how a diocese arrived* at its current portfolio of buildings. Two dioceses could show an identical distribution of present-day redevelopment scores while having gotten there by very different paths — one having steadily right-sized its parish count in step with a declining Catholic population, another having left a 1950s-era building stock largely untouched through decades of demographic change. To add that macro-level context, this section uses a newly assembled long-run diocese-level panel — extracted from digitized editions of *The Official Catholic Directory*, 1888-2021 — to construct a **"parishes per 10,000 Catholics"** ratio over time. A diocese whose ratio has risen over the decades has been closing/consolidating parishes more slowly than its Catholic population has shrunk (i.e., accumulating excess supply); a diocese whose ratio has fallen has consolidated aggressively relative to its own population trend (or its population simply outgrew a roughly fixed parish count).

### Data coverage and reliability caveats

This panel is fundamentally different in character from the Mass-times/Overture/Census data used elsewhere in this paper: it is OCR-derived from century-old print directories, re-extracted in part by an LLM pass over 14 "decade-benchmark" editions (1890, 1900, 1910, 1920, 1930, 1940, 1949, 1960, 1970, 1980, 1990, 2000, 2010, 2020) plus five intermediate gap-fill patches, merged onto a base regex-extraction panel. A few points are worth being explicit about before drawing any conclusions from it:

* **Coverage is heavily skewed toward the recent decade-benchmark years.** Both `total_catholic_population` and (especially) `total_parishes` are sparsely and unevenly populated across the full 1888-2021 span; the 2000/2010/2020 editions have by far the best joint coverage (over 130 dioceses each with both fields present), while 1900, 1910, and 1949 have essentially no usable `total_parishes` values at all in this file.
* **Pre-1930 editions are known to be sparser and less reliable** — fewer dioceses reporting, more OCR noise, and a lower base-panel parse rate generally (the project's own README documents `total_parishes` as a structurally unreliable field, 0.5% raw parse rate, across all eras, only partially improved by the LLM decade re-extraction). We do not treat any pre-1930 point estimate as precise.
* **A validated failure mode exists even in "good" years**: roughly 8.6% of raw `total_catholic_population` values across the whole panel are implausible by magnitude alone (single-year "craters" or OCR repeated-digit "spikes"), and this is only weakly caught by the panel's own quality flags. We therefore apply our own two-stage filter below rather than trusting raw values as-is: (1) absolute plausibility bounds (Catholic population between 1,000 and 6,000,000; parish count between 1 and 1,500), and (2) a within-diocese local-neighborhood swing filter, flagging any single value that deviates by more than 3x (or falls below 1/3) of that diocese's own median value across all its available years within a ±15-year window — the same style of check the panel's own cleaning methodology used. Diocese-year totals are also summed across any duplicate rows sharing a canonical name (e.g., "Anchorage" and "Juneau" before their 2020 administrative union) rather than averaged, to avoid inflating the ratio for sees that were historically split.
* We require at least 4 of the 14 decade-benchmark years to survive this filter for a diocese before including it in the "early vs. late" comparison below, and we compare a multi-year early-period average (1930-1990 benchmarks) against a multi-year late-period average (2000-2020 benchmarks) rather than any single point-to-point pair, to reduce the influence of any one residual bad value.

```{r ocd_panel_load, message=FALSE, warning=FALSE}
library(readr)

decade_years <- c(1890,1900,1910,1920,1930,1940,1949,1960,1970,1980,1990,2000,2010,2020)
early_years  <- c(1930,1940,1949,1960,1970,1980,1990)
late_years   <- c(2000,2010,2020)

ocd_panel <- read_csv("ocd_historical_panel/ocd_diocese_panel_merged.csv", show_col_types = FALSE)

# Stage 1: absolute plausibility bounds, applied row-wise before aggregation
row_ok <- ocd_panel %>%
  filter(!is.na(total_catholic_population) | !is.na(total_parishes)) %>%
  mutate(
    total_catholic_population = if_else(!is.na(total_catholic_population) &
      (total_catholic_population < 1000 | total_catholic_population > 6000000),
      NA_real_, total_catholic_population),
    total_parishes = if_else(!is.na(total_parishes) &
      (total_parishes < 1 | total_parishes > 1500), NA_real_, total_parishes)
  )

# Aggregate to diocese-year (sums across duplicate/pre-merger sees sharing a
# canonical name)
ocd_agg <- row_ok %>%
  group_by(canonical_diocese_name, year) %>%
  summarise(
    total_catholic_population = if (all(is.na(total_catholic_population))) NA_real_ else sum(total_catholic_population, na.rm = TRUE),
    total_parishes = if (all(is.na(total_parishes))) NA_real_ else sum(total_parishes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(total_catholic_population), !is.na(total_parishes),
         total_catholic_population >= 1000, total_parishes >= 1)

# Stage 2: within-diocese local-neighborhood swing filter
flag_outliers <- function(df, field) {
  df %>%
    arrange(canonical_diocese_name, year) %>%
    group_by(canonical_diocese_name) %>%
    mutate(
      local_med = map_dbl(year, function(y) {
        w <- .data[[field]][abs(year - y) <= 15 & year != y]
        if (length(w) == 0) return(NA_real_)
        median(w, na.rm = TRUE)
      }),
      ratio_to_local = .data[[field]] / local_med
    ) %>%
    ungroup()
}

ocd_agg <- flag_outliers(ocd_agg, "total_catholic_population") %>%
  rename(pop_local_med = local_med, pop_ratio = ratio_to_local)
ocd_agg <- flag_outliers(ocd_agg, "total_parishes") %>%
  rename(par_local_med = local_med, par_ratio = ratio_to_local)

ocd_clean <- ocd_agg %>%
  filter(
    is.na(pop_local_med) | (pop_ratio >= 0.3 & pop_ratio <= 3),
    is.na(par_local_med) | (par_ratio >= 0.3 & par_ratio <= 3)
  ) %>%
  mutate(parishes_per_10k = total_parishes / total_catholic_population * 10000)

panel_benchmarks <- ocd_clean %>% filter(year %in% decade_years)

coverage <- panel_benchmarks %>%
  group_by(canonical_diocese_name) %>%
  summarise(n_years = n_distinct(year), .groups = "drop")

early <- panel_benchmarks %>% filter(year %in% early_years) %>%
  group_by(canonical_diocese_name) %>%
  summarise(early_ratio = mean(parishes_per_10k), .groups = "drop")
late <- panel_benchmarks %>% filter(year %in% late_years) %>%
  group_by(canonical_diocese_name) %>%
  summarise(late_ratio = mean(parishes_per_10k), .groups = "drop")

diocese_trend <- inner_join(early, late, by = "canonical_diocese_name") %>%
  left_join(coverage, by = "canonical_diocese_name") %>%
  mutate(pct_change = (late_ratio - early_ratio) / early_ratio * 100) %>%
  filter(n_years >= 4) %>%
  arrange(desc(pct_change))

n_trend_dioceses <- nrow(diocese_trend)
```

After this filtering, `r n_trend_dioceses` dioceses have at least 4 usable decade-benchmark years spanning both an early (1930-1990) and late (2000-2020) period — a meaningful subset of the roughly 195 dioceses that have existed in the modern era, but far from all of them, and concentrated among dioceses whose 1930/1990/2000+ editions happened to parse cleanly. We do not claim national coverage; this is a within-diocese, longitudinal comparison for the subset where the data supports it.

### Which dioceses over-built, and which consolidated aggressively?

```{r ocd_trend_table, message=FALSE, warning=FALSE}
diocese_trend %>%
  arrange(desc(pct_change)) %>%
  mutate(
    early_ratio = round(early_ratio, 2),
    late_ratio = round(late_ratio, 2),
    pct_change = round(pct_change, 1)
  ) %>%
  select(
    Diocese = canonical_diocese_name,
    `Benchmark years used` = n_years,
    `Parishes/10k Catholics (1930-1990 avg.)` = early_ratio,
    `Parishes/10k Catholics (2000-2020 avg.)` = late_ratio,
    `% change` = pct_change
  ) %>%
  reactable(
    defaultPageSize = 10,
    striped = TRUE, highlight = TRUE,
    columns = list(
      `% change` = colDef(style = function(value) {
        color <- if (value > 0) "#D55E00" else "#0072B2"
        list(color = color, fontWeight = "bold")
      })
    )
  )
```

The single most striking finding from this comparison is asymmetry: essentially **no diocese in this filtered sample shows a dramatic *rise* in parishes-per-capita.** The largest increases — Alexandria (Louisiana), Lexington, Las Cruces, Kalamazoo, Houma-Thibodaux — are modest, in the range of +7% to +26%, and several long-established urban archdioceses (New Orleans, Miami, Boston) actually show small *declines* even over this "early" window. By contrast, the largest *declines* in parishes-per-capita are large in magnitude and heavily concentrated among historically rural or "mission-territory" dioceses that industrialized/urbanized rapidly: **Salt Lake City** (-85%), **Wilmington** (-83%), **Saint Augustine** (-82%; Florida), **Green Bay** (-81%), **Amarillo** (-71%), **Anchorage-Juneau** (-71%), and **Santa Fe** (-75%). These dioceses typically had dense networks of small rural/ethnic mission parishes serving a comparatively small early-20th-century Catholic population (e.g., Salt Lake City reported roughly 30 parishes for ~17,000 Catholics in 1930, a ratio of 17.4 per 10,000), and then saw Catholic population grow enormously through 20th-century migration (Salt Lake City's Catholic population is now over 300,000) without anything like a proportional expansion in parish count — the opposite of the "over-building" hypothesis, and a pattern of chronic *under-supply relative to population growth* in the Sun Belt/mission-territory dioceses rather than legacy over-supply in the Rust Belt.

This is a genuinely useful, if unexpected, macro finding for the paper's core thesis: at the aggregate diocese level, and to the extent this filtered historical sample is representative, the dominant long-run pattern among US dioceses is **population growth outrunning parish construction** in the fastest-growing sees, not systematic over-building followed by stranded excess capacity. That does not contradict the parish-level Redevelopment Index — a diocese can simultaneously have too few parishes in its fast-growing suburbs *and* several specific underused, high-parking, low-Mass-frequency urban parishes ripe for conversion — but it does argue against treating "aggregate historical over-supply" as a diocese-wide justification for closures; the case for redevelopment in this paper rests on parish-specific characteristics (Mass frequency, parking footprint, local density), not a diocese having systematically over-built relative to its own history.

### Connecting the historical trend to the Redevelopment Index

Because the historical panel's `canonical_diocese_name` and the paper's own `diocese_name` field are drawn from the same underlying diocese-naming convention, the two datasets merge directly on diocese name (181 of 199 dioceses in the parish-level dataset match a canonical historical diocese; the unmatched remainder are almost entirely Eastern Catholic eparchies/exarchates and the military ordinariate, which the OCD historical panel does not cover on a comparable basis). This lets us test directly whether a diocese's long-run parishes-per-capita trend correlates with its present-day Redevelopment Index scores.

```{r ocd_redev_link, message=FALSE, warning=FALSE}
final_data_flat <- final_data %>% sf::st_drop_geometry()

diocese_redev <- final_data_flat %>%
  filter(!is.na(redev_score_adj)) %>%
  group_by(diocese_name) %>%
  summarise(
    n_parishes_in_data = n(),
    mean_redev_score = mean(redev_score_adj, na.rm = TRUE),
    .groups = "drop"
  )

joined_trend_redev <- diocese_trend %>%
  inner_join(diocese_redev, by = c("canonical_diocese_name" = "diocese_name")) %>%
  filter(n_parishes_in_data >= 5)

cor_test <- cor.test(joined_trend_redev$pct_change, joined_trend_redev$mean_redev_score)
n_joined <- nrow(joined_trend_redev)
r_val <- round(unname(cor_test$estimate), 2)
p_val <- round(cor_test$p.value, 2)
```

Merging the two datasets and requiring at least 5 parishes per diocese in the redevelopment-index data yields `r n_joined` dioceses with both a historical trend and a present-day mean Redevelopment Score. The correlation between a diocese's historical change in parishes-per-10,000-Catholics and its mean present-day Redevelopment Score is **r = `r r_val`** (p = `r p_val`) — weak, in the expected direction (dioceses that consolidated more aggressively historically tend to have slightly *lower* average redevelopment scores today, i.e. less remaining "slack" to redevelop), but **not statistically significant** at this sample size. We report this transparently rather than overstating it: with only `r n_joined` dioceses and the noise inherent in the historical panel, this analysis cannot confirm a robust diocese-level relationship between historical supply trends and present-day parish-level over-capacity. Establishing that link convincingly would require either a larger/cleaner historical panel (extending the LLM re-extraction and gap-filling to more editions and dioceses) or restricting the parish-level Redevelopment Index comparison to the specific decades/dioceses with the strongest historical coverage. We flag this as a natural extension for future work rather than force a conclusion the current data cannot support.

## Policy Implications and Conclusion

The US Catholic Church is currently holding a vast portfolio of economically inefficient assets. By adhering to 19th and 20th-century spatial borders, the institution suffers from both spatial leakage and the immense opportunity costs associated with underutilized urban land.

We propose that Diocesan planners adopt an economically rigorous, data-driven approach to pastoral planning:
1. **Acknowledge Spatial Realities:** Consolidate parishes based on driving-time catchment areas (Voronoi/Isochrone models) rather than historic neighborhood borders.
2. **Monetize the Margins:** Prioritize the conversion of massive surface parking lots (quantified via Overture Maps) in high-density urban zones. Ground-leasing these specific plots for affordable or market-rate housing achieves a dual mandate: providing a stable, highly lucrative revenue stream for the diocese while actively combatting the localized urban housing crisis.

Further research should integrate local municipal zoning codes and historic preservation overlays to calculate the exact Net Present Value (NPV) of these suggested parcel conversions.

## References
*   Alonso, W. (1964). *Location and Land Use*. Harvard University Press.
*   Azzi, C., & Ehrenberg, R. (1975). Household Allocation of Time and Church Attendance. *Journal of Political Economy*, 83(1), 27-56.
*   Hotelling, H. (1929). Stability in Competition. *The Economic Journal*, 39(153), 41-57.
*   Iannaccone, L. R. (1998). Introduction to the Economics of Religion. *Journal of Economic Literature*, 36(3), 1465-1495.
*   Shoup, D. C. (2005). *The High Cost of Free Parking*. American Planning Association.
