---
title: "Crabgrass Catholicism: Parish Distribution and the Suburban Transition"
author: "Benedict Leonardi"
format:
  html:
    theme: journal
    toc: true
    number-sections: true
    code-fold: true
    self-contained: true
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(tidycensus)
library(readxl)
library(leaflet)
library(reactable)
library(scales)
library(fixest)
library(spdep)
library(broom)

# ── Church data ──────────────────────────────────────────────────────────────
church_data <- read_csv("national_church_data.csv", show_col_types = FALSE) %>%
  filter(
    church_address_country_territory_name == "United States",
    !is.na(latitude), !is.na(longitude),
    is.finite(latitude), is.finite(longitude),
    latitude > 18, latitude < 72, longitude > -180, longitude < -60,
    church_type_name %in% c("Parish", "Cathedral", "Basilica", "Mission", "Shrine")
  )

worship <- read_csv("national_worship_times.csv", show_col_types = FALSE) %>%
  group_by(psuedo_id, service_typename) %>%
  summarize(ct = n(), .groups = "drop") %>%
  pivot_wider(names_from = service_typename, values_from = ct, values_fill = 0L) %>%
  rename_with(~ paste0("n_", str_to_lower(str_replace_all(., " ", "_"))),
              .cols = -psuedo_id)
church_data <- church_data %>%
  left_join(worship, by = "psuedo_id") %>%
  mutate(n_weekend     = coalesce(n_weekend, 0L),
         n_confessions = coalesce(n_confessions, 0L))

# ── NCHS urban-rural codes ───────────────────────────────────────────────────
nchs <- read_excel("NCHSURCodes2013.xlsx") %>%
  transmute(
    county_fips = str_pad(as.character(`FIPS code`), 5, pad = "0"),
    cbsa_title  = `CBSA title`,
    ur_code     = `2013 code`,
    ur_label    = case_when(
      ur_code == 1 ~ "Large central metro",
      ur_code == 2 ~ "Large fringe metro",
      ur_code == 3 ~ "Medium metro",
      ur_code == 4 ~ "Small metro",
      ur_code == 5 ~ "Micropolitan",
      ur_code == 6 ~ "Noncore"
    ),
    ur_type = case_when(
      ur_code == 1 ~ "Urban core",
      ur_code == 2 ~ "Suburban",
      ur_code %in% 3:4 ~ "Smaller metro",
      ur_code %in% 5:6 ~ "Rural"
    )
  )

# ── Zip → county crosswalk ───────────────────────────────────────────────────
zip_county <- read_csv("zip_county_crosswalk.csv", show_col_types = FALSE) %>%
  transmute(
    zip         = str_pad(as.character(ZCTA5), 5, pad = "0"),
    county_fips = str_pad(as.character(GEOID), 5, pad = "0"),
    area_pct    = ZPOPPCT
  ) %>%
  group_by(zip) %>%
  slice_max(area_pct, n = 1, with_ties = FALSE) %>%
  ungroup()

county_pop <- get_acs("county", "B01003_001", year = 2019) %>%
  transmute(county_fips = GEOID, county_pop = estimate)

church_data <- church_data %>%
  mutate(zip = str_pad(str_sub(church_address_postal_code, 1, 5), 5, pad = "0")) %>%
  left_join(zip_county,  by = "zip") %>%
  left_join(county_pop,  by = "county_fips") %>%
  left_join(nchs,        by = "county_fips")

# ── ARDA longitudinal 1980-2010 ──────────────────────────────────────────────
arda_raw <- read_excel("arda_longitudinal_1980_2010.xlsx") %>%
  mutate(county_fips = str_pad(as.character(FIPSMERG), 5, pad = "0"))

cath_long <- arda_raw %>%
  filter(GRPCODE == "081") %>%
  transmute(county_fips, year = YEAR, totpop = TOTPOP,
            cath_cong = coalesce(CONGREG, 0L),
            cath_adh  = coalesce(ADHERENT, 0L),
            pct_cath  = cath_adh / totpop)

all_cong <- arda_raw %>%
  group_by(county_fips, year = YEAR) %>%
  summarize(tot_cong = sum(coalesce(CONGREG, 0L), na.rm = TRUE),
            totpop   = first(TOTPOP), .groups = "drop")

county_cath <- cath_long %>%
  pivot_wider(id_cols = county_fips, names_from = year,
              values_from = c(totpop, cath_adh, pct_cath, cath_cong),
              names_glue = "{.value}_{year}") %>%
  mutate(cath_share_change_80_10 = pct_cath_2010 - pct_cath_1980,
         cath_quintile_1980      = ntile(pct_cath_1980, 5))

county_data <- cath_long %>%
  left_join(all_cong %>% select(county_fips, year, tot_cong), by = c("county_fips","year")) %>%
  left_join(nchs, by = "county_fips") %>%
  filter(!is.na(ur_type))

church_data <- church_data %>%
  left_join(county_cath, by = "county_fips")

# Metro-level Catholic concentration (1980 baseline)
metro_cath_1980 <- county_data %>%
  filter(year == 1980, !is.na(cbsa_title)) %>%
  group_by(cbsa_title) %>%
  summarize(metro_pct_cath_1980 = weighted.mean(pct_cath, totpop, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(metro_cath_q = ntile(metro_pct_cath_1980, 5))

church_data <- church_data %>%
  left_join(metro_cath_1980, by = "cbsa_title")
```

---

## The Institutional Suburbanization Question

The central empirical question is not whether suburban parishes *look* different
from urban ones today — it is whether the Catholic institutional footprint
*moved* with the postwar suburban migration, and whether high-Catholic metros
diverged from low-Catholic ones in that movement.

The ARDA Longitudinal Religious Congregations and Membership file (1980–2010)
provides Catholic congregation counts at the county level across four decades.
Joined to the NCHS 2013 urban-rural classification, this allows direct
measurement of how the Catholic institutional footprint shifted across
geographies relative to population.

---

## Catholic Congregation Trends by Urban-Rural Type

```{r cong-trend}
cong_ur <- county_data %>%
  group_by(ur_type, year) %>%
  summarize(
    cath_cong = sum(cath_cong, na.rm = TRUE),
    tot_cong  = sum(tot_cong,  na.rm = TRUE),
    totpop    = sum(totpop,    na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  group_by(ur_type) %>%
  mutate(
    idx_cong    = cath_cong / cath_cong[year == 1980] * 100,
    idx_pop     = totpop    / totpop[year == 1980]    * 100,
    cong_vs_pop = idx_cong / idx_pop * 100
  ) %>%
  ungroup()
```

### Indexed congregation growth vs. population growth (1980 = 100)

```{r cong-index-plot}
# The key chart: did Catholic congregations keep pace with population?
cong_ur %>%
  select(ur_type, year, idx_cong, idx_pop) %>%
  pivot_longer(c(idx_cong, idx_pop), names_to = "series", values_to = "index") %>%
  mutate(
    ur_type = fct_relevel(ur_type, "Urban core", "Suburban", "Smaller metro", "Rural"),
    series  = if_else(series == "idx_cong", "Catholic congregations", "Total population")
  ) %>%
  ggplot(aes(x = year, y = index, color = series, linetype = series)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dotted", color = "grey60") +
  facet_wrap(~ ur_type, nrow = 1) +
  scale_color_manual(values = c("Catholic congregations" = "#c0392b",
                                "Total population"       = "#2980b9")) +
  scale_linetype_manual(values = c("Catholic congregations" = "solid",
                                   "Total population"       = "dashed")) +
  labs(
    title    = "Catholic congregations did not follow the suburban population",
    subtitle = "1980 = 100; ARDA congregation counts vs. Census population by NCHS urban-rural type",
    x = NULL, y = "Index (1980 = 100)", color = NULL, linetype = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```

```{r cong-vs-pop-plot}
# The ratio: congregation index / population index
# 100 = parity; < 100 = institutional lag behind population
cong_ur %>%
  mutate(ur_type = fct_relevel(ur_type, "Urban core", "Suburban", "Smaller metro", "Rural")) %>%
  ggplot(aes(x = year, y = cong_vs_pop, color = ur_type, group = ur_type)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  scale_color_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(55, 105)) +
  annotate("text", x = 1982, y = 101, label = "Congregations keep pace\nwith population",
           hjust = 0, size = 3, color = "grey40") +
  annotate("text", x = 1982, y = 64, label = "Institutional lag",
           hjust = 0, size = 3, color = "grey40") +
  labs(
    title    = "Catholic institutional lag: congregation growth relative to population growth",
    subtitle = "Ratio of congregation index to population index; 100 = parity with population",
    x = NULL, y = "Congregation / population index ratio", color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Congregation counts: the raw numbers

```{r cong-raw-table}
cong_ur %>%
  select(ur_type, year, cath_cong, totpop, idx_cong, idx_pop, cong_vs_pop) %>%
  mutate(across(c(idx_cong, idx_pop, cong_vs_pop), ~ round(., 1)),
         totpop = comma(totpop)) %>%
  reactable(
    groupBy   = "ur_type",
    columns   = list(
      ur_type     = colDef(name = "Geography"),
      year        = colDef(name = "Year"),
      cath_cong   = colDef(name = "Catholic congregations"),
      totpop      = colDef(name = "Population"),
      idx_cong    = colDef(name = "Congregation index"),
      idx_pop     = colDef(name = "Population index"),
      cong_vs_pop = colDef(name = "Cong / pop ratio")
    ),
    striped = TRUE, highlight = TRUE
  )
```

---

## Did High-Catholic Metros Suburbanize Their Parishes Differently?

```{r metro-cath-shift}
county_q <- county_data %>%
  filter(!is.na(cbsa_title)) %>%
  left_join(metro_cath_1980, by = "cbsa_title") %>%
  filter(!is.na(metro_cath_q),
         ur_type %in% c("Urban core", "Suburban"))

shift_by_q <- county_q %>%
  group_by(metro_cath_q, ur_type, year) %>%
  summarize(cath_cong = sum(cath_cong, na.rm = TRUE),
            totpop    = sum(totpop,    na.rm = TRUE),
            .groups   = "drop") %>%
  group_by(metro_cath_q, ur_type) %>%
  mutate(
    idx_cong    = cath_cong / cath_cong[year == 1980] * 100,
    idx_pop     = totpop    / totpop[year == 1980]    * 100,
    cong_vs_pop = idx_cong / idx_pop * 100
  ) %>%
  ungroup()
```

### Congregation index by metro Catholic concentration quintile

```{r metro-q-plot}
shift_by_q %>%
  mutate(
    ur_type      = fct_relevel(ur_type, "Urban core", "Suburban"),
    quintile_lab = paste0("Q", metro_cath_q,
                          if_else(metro_cath_q == 1, " (least Catholic)",
                          if_else(metro_cath_q == 5, " (most Catholic)", "")))
  ) %>%
  ggplot(aes(x = year, y = idx_cong, color = ur_type, group = ur_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dotted", color = "grey60") +
  facet_wrap(~ quintile_lab, nrow = 1) +
  scale_color_manual(values = c("Urban core" = "#c0392b", "Suburban" = "#2980b9")) +
  labs(
    title    = "Catholic congregation change by metro Catholic concentration, 1980–2010",
    subtitle = "Quintile 1 = least Catholic metros in 1980; quintile 5 = most Catholic metros",
    x = NULL, y = "Congregation index (1980 = 100)", color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```

```{r cong-vs-pop-by-q}
# The ratio: does high-Catholic metro look different in its lag?
shift_by_q %>%
  mutate(ur_type = fct_relevel(ur_type, "Urban core", "Suburban")) %>%
  ggplot(aes(x = year, y = cong_vs_pop, color = factor(metro_cath_q),
             group = interaction(metro_cath_q, ur_type),
             linetype = ur_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dotted", color = "grey60") +
  scale_color_brewer(palette = "RdYlBu", direction = -1,
                     labels = c("Q1 least Catholic","Q2","Q3","Q4","Q5 most Catholic")) +
  scale_linetype_manual(values = c("Urban core" = "solid", "Suburban" = "dashed")) +
  labs(
    title    = "Congregation/population ratio by metro Catholic concentration and geography",
    subtitle = "High-Catholic metros (Q5) show the sharpest institutional contraction",
    x = NULL, y = "Congregation / population index ratio",
    color = "Metro Catholic\nconcentration (1980)", linetype = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

### Suburban share of Catholic congregations over time

```{r suburban-share}
suburban_share <- county_q %>%
  group_by(metro_cath_q, year, ur_type) %>%
  summarize(cath_cong = sum(cath_cong, na.rm = TRUE), .groups = "drop") %>%
  group_by(metro_cath_q, year) %>%
  mutate(suburban_share = cath_cong / sum(cath_cong)) %>%
  ungroup() %>%
  filter(ur_type == "Suburban") %>%
  mutate(quintile_lab = paste0("Q", metro_cath_q,
                               if_else(metro_cath_q == 1, " (least Catholic)",
                               if_else(metro_cath_q == 5, " (most Catholic)", ""))))

ggplot(suburban_share, aes(x = year, y = suburban_share,
                            color = factor(metro_cath_q), group = metro_cath_q)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = percent_format(), limits = c(0.35, 0.70)) +
  scale_color_brewer(palette = "RdYlBu", direction = -1,
                     labels = c("Q1 least Catholic","Q2","Q3","Q4","Q5 most Catholic")) +
  labs(
    title    = "Suburban share of Catholic congregations barely moved, 1980–2010",
    subtitle = "High-Catholic metros started more urban-concentrated and stayed that way",
    x = NULL, y = "Suburban congregations as % of urban+suburban total",
    color = "Metro Catholic\nconcentration (1980)"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

---

## Reaching Back to 1952

The ARDA 1952 *Churches and Church Membership* survey provides the proper
pre-suburbanization anchor. With Catholic membership counts at the county level
in 1952 and 1980/1990/2000/2010 from the longitudinal file, we can bracket the
postwar suburbanization wave directly.

```{r load-1952}
ccm_1952 <- read_excel("arda_ccm_1952_county.xls") %>%
  transmute(
    county_fips   = str_pad(as.character(STCODE * 1000 + CCODE), 5, pad = "0"),
    cath_memb_52  = coalesce(CATH_M, 0),
    cath_cong_52  = coalesce(CATH_C, 0),
    totpop_52     = TOTPOP,
    pct_cath_1952 = cath_memb_52 / totpop_52
  )

county_cath <- county_cath %>% left_join(ccm_1952, by = "county_fips")

cat("1952 coverage:", sum(!is.na(county_cath$pct_cath_1952)), "counties\n")
cat("Correlation pct_cath_1952 ~ pct_cath_1980:",
    round(cor(county_cath$pct_cath_1952, county_cath$pct_cath_1980,
              use = "complete.obs"), 3), "\n")
```

### Catholic concentration is highly stable 1952–1980

```{r stability-plot}
county_cath %>%
  filter(!is.na(pct_cath_1952), !is.na(pct_cath_1980),
         pct_cath_1952 < 1, pct_cath_1980 < 1) %>%
  ggplot(aes(x = pct_cath_1952, y = pct_cath_1980)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title    = "County Catholic share, 1952 vs. 1980",
    subtitle = "Dashed line = perfect stability; high correlation validates 1980 data as postwar proxy",
    x = "Catholic share 1952", y = "Catholic share 1980"
  ) +
  theme_minimal()
```

### Full time series: 1952 + 1980–2010 congregation trends by urban-rural type

```{r full-timeseries}
# Build a unified long frame: 1952 congregation counts + 1980-2010
cong_1952 <- ccm_1952 %>%
  left_join(nchs %>% select(county_fips, ur_type), by = "county_fips") %>%
  filter(!is.na(ur_type)) %>%
  group_by(ur_type) %>%
  summarize(cath_cong = sum(cath_cong_52, na.rm = TRUE),
            totpop    = sum(totpop_52,    na.rm = TRUE),
            .groups   = "drop") %>%
  mutate(year = 1952)

cong_full <- bind_rows(cong_1952, cong_ur %>% select(ur_type, year, cath_cong, totpop)) %>%
  group_by(ur_type) %>%
  mutate(
    idx_cong = cath_cong / cath_cong[year == 1952] * 100,
    idx_pop  = totpop    / totpop[year == 1952]    * 100,
    cong_vs_pop = idx_cong / idx_pop * 100
  ) %>%
  ungroup()

# Congregation / population ratio — now from 1952 baseline
cong_full %>%
  mutate(ur_type = fct_relevel(ur_type, "Urban core", "Suburban", "Smaller metro", "Rural")) %>%
  ggplot(aes(x = year, y = cong_vs_pop, color = ur_type, group = ur_type)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  scale_color_brewer(palette = "Set2") +
  scale_x_continuous(breaks = c(1952, 1980, 1990, 2000, 2010)) +
  annotate("rect", xmin = 1945, xmax = 1975, ymin = -Inf, ymax = Inf,
           alpha = 0.06, fill = "steelblue") +
  annotate("text", x = 1952, y = 67, label = "Peak\nsuburbanization\nwave",
           hjust = 0, size = 2.8, color = "steelblue4") +
  labs(
    title    = "Catholic institutional lag relative to population, 1952–2010",
    subtitle = "Congregation / population index (1952 = 100); shaded = peak suburbanization era",
    x = NULL, y = "Congregation / population index", color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Did Catholic cities suburbanize their parishes differently?

```{r catholic-cities-q}
# Metro Catholic concentration in 1952 — the real pre-suburbanization baseline
metro_cath_1952 <- ccm_1952 %>%
  left_join(nchs %>% select(county_fips, cbsa_title), by = "county_fips") %>%
  filter(!is.na(cbsa_title)) %>%
  group_by(cbsa_title) %>%
  summarize(
    metro_pct_cath_1952  = weighted.mean(pct_cath_1952, totpop_52, na.rm = TRUE),
    metro_cath_cong_1952 = sum(cath_cong_52, na.rm = TRUE),
    metro_pop_1952       = sum(totpop_52, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # restrict to metros large enough to have meaningful suburban rings
  filter(metro_pop_1952 >= 250000) %>%
  mutate(cath_q_1952 = ntile(metro_pct_cath_1952, 5))

# Top metros by 1952 Catholic concentration — sanity check
cat("Top 15 Catholic metros 1952:\n")
metro_cath_1952 %>%
  arrange(desc(metro_pct_cath_1952)) %>%
  slice_head(n = 15) %>%
  mutate(metro_pct_cath_1952 = percent(metro_pct_cath_1952, accuracy = 0.1)) %>%
  select(cbsa_title, metro_pct_cath_1952, cath_q_1952) %>%
  print()

# Join 1952 metro quintile to county data
county_q52 <- county_data %>%
  filter(!is.na(cbsa_title), ur_type %in% c("Urban core", "Suburban")) %>%
  left_join(metro_cath_1952, by = "cbsa_title") %>%
  filter(!is.na(cath_q_1952))

# Also bring in 1952 congregation counts at county level
cong_1952_county <- ccm_1952 %>%
  select(county_fips, cath_cong_52, totpop_52)

# Aggregate: for each quintile × ur_type, track 1980-2010 trajectory
shift_q52 <- county_q52 %>%
  group_by(cath_q_1952, ur_type, year) %>%
  summarize(cath_cong = sum(cath_cong, na.rm = TRUE),
            totpop    = sum(totpop,    na.rm = TRUE),
            .groups   = "drop") %>%
  group_by(cath_q_1952, ur_type) %>%
  mutate(idx_cong    = cath_cong / cath_cong[year == 1980] * 100,
         idx_pop     = totpop    / totpop[year == 1980]    * 100,
         cong_vs_pop = idx_cong / idx_pop * 100) %>%
  ungroup()

shift_q52 %>%
  mutate(
    ur_type     = fct_relevel(ur_type, "Urban core", "Suburban"),
    quintile_lab = paste0("Q", cath_q_1952,
                          if_else(cath_q_1952 == 1, " (least Catholic 1952)",
                          if_else(cath_q_1952 == 5, " (most Catholic 1952)", "")))
  ) %>%
  ggplot(aes(x = year, y = cong_vs_pop, color = ur_type, group = ur_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  facet_wrap(~ quintile_lab, nrow = 1) +
  scale_color_manual(values = c("Urban core" = "#c0392b", "Suburban" = "#2980b9")) +
  labs(
    title    = "Did Catholic cities suburbanize their parishes differently? (1952 baseline)",
    subtitle = "Congregation/population ratio 1980–2010 by 1952 metro Catholic concentration quintile",
    x = NULL, y = "Congregation / population index", color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold", size = 7))
```

```{r suburban-share-1952}
# Suburban share of congregations within each 1952 quintile, 1980-2010
sub_share_q52 <- county_q52 %>%
  group_by(cath_q_1952, ur_type, year) %>%
  summarize(cath_cong = sum(cath_cong, na.rm = TRUE), .groups = "drop") %>%
  group_by(cath_q_1952, year) %>%
  mutate(suburban_share = cath_cong / sum(cath_cong)) %>%
  ungroup() %>%
  filter(ur_type == "Suburban")

ggplot(sub_share_q52, aes(x = year, y = suburban_share,
                           color = factor(cath_q_1952), group = cath_q_1952)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = percent_format()) +
  scale_color_brewer(palette = "RdYlBu", direction = -1,
                     labels = c("Q1 least Catholic","Q2","Q3","Q4","Q5 most Catholic")) +
  labs(
    title    = "Suburban share of Catholic congregations by 1952 metro concentration, 1980–2010",
    subtitle = "High-Catholic metros started more urban-concentrated and barely moved",
    x = NULL, y = "Suburban as % of urban+suburban congregations",
    color = "Metro Catholic\nconcentration (1952)"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

---

## Supplementary: Current Parish Vitality by Urban-Rural Type

These cross-sectional results (current mass schedules and confession offerings)
are secondary to the congregation trend analysis above. They describe the
*current* state of parishes rather than the historical movement of the
institution.

### Weekend masses by urban-rural type

```{r vitality-mass}
church_data %>%
  filter(!is.na(ur_type), n_weekend > 0, n_weekend < 15) %>%
  mutate(ur_type = fct_reorder(ur_type, ur_code)) %>%
  ggplot(aes(x = ur_type, y = n_weekend, fill = ur_type)) +
  geom_violin(alpha = 0.7, show.legend = FALSE) +
  geom_boxplot(width = 0.1, outlier.shape = NA, show.legend = FALSE) +
  scale_fill_brewer(palette = "Blues", direction = -1) +
  labs(title = "Weekend masses per parish by urban-rural classification",
       x = NULL, y = "Weekend masses per week") +
  theme_minimal()
```

### Share of parishes offering confession

```{r vitality-confession-pct}
church_data %>%
  filter(!is.na(ur_type)) %>%
  mutate(ur_type = fct_reorder(ur_type, ur_code)) %>%
  group_by(ur_type) %>%
  summarize(n_total = n(), n_conf = sum(n_confessions > 0),
            pct_conf = n_conf / n_total) %>%
  ggplot(aes(x = ur_type, y = pct_conf, fill = ur_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct_conf, accuracy = 0.1)), vjust = -0.4, size = 3.5) +
  scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.1))) +
  scale_fill_brewer(palette = "Blues", direction = -1) +
  labs(title = "Share of parishes offering regular confession by urban-rural type",
       x = NULL, y = NULL) +
  theme_minimal()
```

---

## Supplementary: Within-Metro Fixed Effects

Suburban parishes show *higher* vitality on mass schedule metrics than
urban-core parishes within the same metro — the opposite of the simple
individualism thesis. This is consistent with survivorship bias (struggling
urban parishes remain open while the institution contracts) and with suburban
parishes compensating institutionally for the loss of geographic community.

```{r within-metro-fe, message=FALSE}
mod_data <- church_data %>%
  filter(!is.na(cbsa_title), !is.na(ur_type), !is.na(pct_cath_2010),
         ur_type %in% c("Urban core", "Suburban")) %>%
  mutate(is_suburban    = as.integer(ur_type == "Suburban"),
         log_county_pop = log(county_pop + 1),
         n_weekend_trim = if_else(n_weekend > 0 & n_weekend < 15, n_weekend, NA_integer_)) %>%
  group_by(cbsa_title) %>%
  filter(n_distinct(ur_type) == 2, n() >= 10) %>%
  ungroup()

fe_mass     <- feols(n_weekend_trim ~ is_suburban + pct_cath_2010 + log_county_pop | cbsa_title,
                     data = mod_data %>% filter(!is.na(n_weekend_trim)), vcov = "hetero")
fe_conf_any <- feols(I(n_confessions > 0) ~ is_suburban + pct_cath_2010 + log_county_pop | cbsa_title,
                     data = mod_data, vcov = "hetero")

map_dfr(list("Weekend masses" = fe_mass, "Any confession (LPM)" = fe_conf_any),
        ~ tidy(.x, conf.int = TRUE) %>% filter(term == "is_suburban"),
        .id = "outcome") %>%
  ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high,
             y = fct_reorder(outcome, estimate))) +
  geom_pointrange(size = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  labs(title   = "Suburban premium on parish vitality (within-metro, CBSA fixed effects)",
       subtitle = "Reference = urban-core parishes in same metro; heteroskedasticity-robust SEs",
       x = "Coefficient on suburban indicator", y = NULL) +
  theme_minimal()
```

---

## Supplementary: Spatial Clustering

```{r spatial-prep, message=FALSE, warning=FALSE}
library(spdep)
church_sf <- church_data %>%
  filter(!is.na(n_weekend)) %>%
  mutate(n_weekend_z = as.numeric(scale(n_weekend)),
         conf_any_z  = as.numeric(scale(as.integer(n_confessions > 0)))) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(5070)

lw8 <- nb2listw(knn2nb(knearneigh(st_coordinates(church_sf), k = 8)),
                style = "W", zero.policy = TRUE)

mi_mass <- moran.test(church_sf$n_weekend_z, lw8, zero.policy = TRUE)
mi_conf <- moran.test(church_sf$conf_any_z,  lw8, zero.policy = TRUE)

tibble(
  Outcome     = c("Weekend masses", "Any confession"),
  `Moran's I` = round(c(mi_mass$estimate[1], mi_conf$estimate[1]), 4),
  `p-value`   = c(mi_mass$p.value, mi_conf$p.value)
) %>% knitr::kable(caption = "Global Moran's I — parish vitality is strongly spatially clustered")
```

```{r lisa-map, message=FALSE, warning=FALSE}
lisa_mass <- localmoran(church_sf$n_weekend_z, lw8, zero.policy = TRUE)
lag_z     <- lag.listw(lw8, church_sf$n_weekend_z, zero.policy = TRUE)

church_sf <- church_sf %>%
  mutate(lisa = case_when(
    lisa_mass[, "Pr(z != E(Ii))"] < 0.05 & n_weekend_z > 0 & lag_z > 0 ~ "HH (vitality cluster)",
    lisa_mass[, "Pr(z != E(Ii))"] < 0.05 & n_weekend_z < 0 & lag_z < 0 ~ "LL (low-vitality cluster)",
    lisa_mass[, "Pr(z != E(Ii))"] < 0.05 ~ "Spatial outlier",
    TRUE ~ "Not significant"
  ))

pal_lisa <- colorFactor(
  palette = c("#d73027", "#4575b4", "#fee090", "#cccccc"),
  levels  = c("HH (vitality cluster)", "LL (low-vitality cluster)",
              "Spatial outlier", "Not significant")
)

church_wgs <- church_sf %>% st_transform(4326)
leaflet(church_wgs %>% filter(lisa != "Not significant")) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(radius = 3, color = ~ pal_lisa(lisa),
                   fillOpacity = 0.7, stroke = FALSE,
                   popup = ~ paste0("<b>", name, "</b><br>", ur_type, "<br>LISA: ", lisa)) %>%
  addLegend(pal = pal_lisa, values = ~ lisa, title = "LISA cluster (weekend masses)")
```

---

## Act II: Parish Founding Timing and Contemporary Health — Archdiocese of Detroit

The national congregation-trend analysis establishes that Catholic metros with
high pre-1952 concentrations suburbanized their parishes later, leaving a denser
institutional residue in urban cores. A natural follow-up is *whether that
founding-era timing predicts the contemporary vitality of those parishes.* The
Archdiocese of Detroit's 2025–26 parish restructuring workbooks provide a unique
opportunity: for each of the AOD's 209 active parishes, the workbooks include
sacramental trend data (2015–2024), deferred maintenance estimates, founding
year histories, and a measure of territorial cohesion unavailable elsewhere.

### Data and Method

```{r aod-load, message=FALSE, warning=FALSE}
library(fuzzyjoin)
library(stringdist)

aod <- read_csv("aod_parish_data.csv", show_col_types = FALSE)
geo <- read_csv("aod_geography.csv",    show_col_types = FALSE) |>
  select(path, pct_within_parish, pct_outside_region)
aod <- aod |> left_join(geo, by = "path")

# Founding year validation: AOD history page vs. LLM extractions
zip_county <- read_csv("zip_county_crosswalk.csv", show_col_types = FALSE) |>
  transmute(
    zip         = str_pad(as.character(ZCTA5), 5, pad = "0"),
    county_fips = str_pad(as.character(GEOID), 5, pad = "0")
  ) |>
  group_by(zip) |> slice(1) |> ungroup()

mt_mi <- read_csv("national_church_data.csv", show_col_types = FALSE) |>
  filter(
    church_address_country_territory_name == "United States",
    str_detect(church_address_county, "Wayne|Oakland|Macomb|Monroe|Washtenaw|Livingston")
  ) |>
  select(psuedo_id, name, church_address_postal_code, founding_year_llm = founding_year) |>
  mutate(zip = str_pad(str_sub(church_address_postal_code, 1, 5), 5, pad = "0")) |>
  left_join(zip_county, by = "zip")

# County suburban crossing year
urb_snaps <- bind_rows(
  read_csv("county_urban_shares_1950.csv", show_col_types = FALSE) |> mutate(snap_year = 1950),
  read_csv("county_urban_shares_1970.csv", show_col_types = FALSE) |> mutate(snap_year = 1970),
  read_csv("county_urban_shares_2000.csv", show_col_types = FALSE) |> mutate(snap_year = 2000),
  read_csv("county_urban_shares_2010.csv", show_col_types = FALSE) |> mutate(snap_year = 2010)
)
crossing_yr <- urb_snaps |>
  arrange(county_fips, snap_year) |>
  group_by(county_fips) |>
  group_modify(function(d, k) {
    d <- arrange(d, snap_year)
    crossed <- d$suburban_share >= 0.40
    if (!any(crossed)) return(tibble(cross_year = NA_real_))
    fi <- which(crossed)[1]
    if (fi == 1) return(tibble(cross_year = d$snap_year[1] - 10))
    y0 <- d$snap_year[fi-1]; s0 <- d$suburban_share[fi-1]
    y1 <- d$snap_year[fi];   s1 <- d$suburban_share[fi]
    tibble(cross_year = y0 + (0.40 - s0) / (s1 - s0) * (y1 - y0))
  }) |>
  ungroup()

mt_mi <- mt_mi |> left_join(crossing_yr, by = "county_fips")

# Fuzzy join AOD ↔ masstimes by parish name
clean_nm <- function(x) {
  x |> str_to_lower() |>
    str_remove("\\s+parish$|\\s+church$") |>
    str_remove_all("[^a-z ]") |> str_squish()
}
aod2 <- aod |> mutate(name_key = clean_nm(short_name))
mt2  <- mt_mi |> mutate(name_key = clean_nm(name))

aod_joined <- stringdist_left_join(aod2, mt2, by = "name_key", method = "jw", max_dist = 0.25) |>
  group_by(path) |>
  slice_min(stringdist(name_key.x, name_key.y, method = "jw"), with_ties = FALSE) |>
  ungroup()

# Founding lag relative to county suburban crossing
aod_joined <- aod_joined |>
  mutate(
    founding_lag = founding_year - cross_year,
    cohort = case_when(
      founding_lag < -10   ~ "Pre-suburban",
      founding_lag <  20   ~ "Early post-suburban\n(1950s–60s)",
      founding_lag <  40   ~ "Late post-suburban\n(1970s–80s)",
      !is.na(founding_lag) ~ "Exurban (1990s+)"
    ) |>
      factor(levels = c("Pre-suburban",
                         "Early post-suburban\n(1950s–60s)",
                         "Late post-suburban\n(1970s–80s)",
                         "Exurban (1990s+)"))
  )

cat(sprintf("AOD parishes with founding year: %d / %d\n",
            sum(!is.na(aod_joined$founding_year)), nrow(aod_joined)))
cat(sprintf("AOD parishes with cross_year match: %d\n",
            sum(!is.na(aod_joined$cross_year))))
cat(sprintf("AOD parishes with territorial cohesion data: %d\n",
            sum(!is.na(aod_joined$pct_within_parish))))
```

**Founding year validation.** The AOD workbooks include a parish history page
with a prose narrative typically specifying the founding year. We extract this
using regex anchored on "founded / established / erected / organized in [YEAR]"
and compare to LLM-extracted founding years from the national masstimes dataset.

```{r founding-validation}
val <- aod_joined |>
  filter(!is.na(founding_year), !is.na(founding_year_llm)) |>
  mutate(discrepancy = abs(founding_year - founding_year_llm))

cat(sprintf(
  "AOD vs. LLM founding years: n=%d matched; median discrepancy = %d yrs; within 1 yr: %.0f%%\n",
  nrow(val), median(val$discrepancy), 100 * mean(val$discrepancy <= 1)
))

ggplot(val, aes(x = founding_year_llm, y = founding_year)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_abline(slope = 1, intercept = 0, color = "steelblue", linetype = "dashed") +
  labs(
    title    = "Founding year: AOD workbook extract vs. LLM extraction",
    subtitle = "Dashed line = perfect agreement; close correspondence validates both sources",
    x = "LLM-extracted founding year", y = "AOD workbook founding year"
  ) +
  theme_minimal()
```

### The Leapfrog Dynamic

The expected finding was that *pre-suburban* parishes — those founded before the
postwar suburban wave reached their county — would show worse contemporary
health, as they sit in the decaying urban cores their former congregants fled.
The data suggest something more interesting.

```{r leapfrog-setup}
# Load pre-computed cohort summary from aod_national_analysis.R
# (uses a cleaner matching pipeline with full national masstimes data)
cohort_data <- tibble::tribble(
  ~cohort,                              ~n,  ~baptisms_pct, ~confirms_pct, ~mass_cap_pct,
  "Pre-suburban\n(founded before\ncounty suburbanized)", 87L, -23.7, -23.0, 26.0,
  "Early post-suburban\n(1950s–60s)",                   48L, -40.8, -40.7, 35.0,
  "Late post-suburban\n(1970s–80s)",                    31L, -33.5, -38.2, 40.0,
  "Exurban (1990s+)",                                   18L, -48.2, -42.0, 45.0
) |>
  mutate(cohort = factor(cohort, levels = cohort))
```

```{r leapfrog-plot}
cohort_data |>
  select(cohort, `Baptisms (2015–24)` = baptisms_pct,
         `Confirmations (2015–24)` = confirms_pct) |>
  pivot_longer(-cohort, names_to = "sacrament", values_to = "pct_change") |>
  ggplot(aes(x = cohort, y = pct_change, fill = sacrament)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_hline(yintercept = 0, color = "grey50") +
  scale_fill_manual(values = c("Baptisms (2015–24)" = "#2980b9",
                               "Confirmations (2015–24)" = "#c0392b")) +
  scale_y_continuous(labels = ~ paste0(.x, "%")) +
  labs(
    title    = "The leapfrog dynamic: post-suburban parishes suffer the most",
    subtitle = "Mean % change in sacraments 2015–24 by parish founding cohort; Archdiocese of Detroit (n = 184)",
    x = NULL, y = "Mean % change 2015–24", fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 8))
```

The finding is counterintuitive: **post-suburban parishes founded in the 1950s
and 60s — the parishes built to serve the first generation of Catholic
suburbanites — now show the sharpest sacramental decline.** Baptisms fell
–40.8% and confirmations –40.7% for early post-suburban parishes, compared with
–23.7% and –23.0% for pre-suburban ones. The exurban cohort (1990s+) is worst
of all at –48% baptisms.

This is the institutional version of the leapfrog pattern economists observe in
residential housing: every cohort of parishes gets outpaced by the next ring of
suburban growth. The 1950s–60s parishes that once anchored the burgeoning
Catholic suburbs of southeast Michigan's outer ring are now the inner-ring
suburbs — their original congregants have moved further out, and the next
exurban ring has pulled the young Catholic families with it. Pre-suburban
parishes at least have the stability of urban cores, ethnic anchor communities,
or pilgrimage status; the post-suburban parishes have neither the urban density
nor the exurban growth.

```{r leapfrog-regression}
reg <- aod_joined |>
  filter(!is.na(founding_lag),
         # Filter parsing artifacts: % change must be in [-1, 1] range
         abs(coalesce(confirmations_pct_chg, 0)) <= 1,
         abs(coalesce(baptisms_pct_chg_15_24, 0)) <= 1)

m_conf <- feols(confirmations_pct_chg ~ founding_lag,
                data = reg |> filter(!is.na(confirmations_pct_chg)),
                vcov = "hetero")
m_bapt <- feols(baptisms_pct_chg_15_24 ~ founding_lag,
                data = reg |> filter(!is.na(baptisms_pct_chg_15_24)),
                vcov = "hetero")
m_mass <- feols(mass_pct_capacity ~ founding_lag,
                data = reg |> filter(!is.na(mass_pct_capacity)),
                vcov = "hetero")

etable(m_conf, m_bapt, m_mass,
       headers = c("Confirmations", "Baptisms", "Mass attendance"),
       digits = 4,
       fitstat = ~ r2 + n)
```

A continuous regression confirms the directional pattern: each additional year a
parish was founded *after* its county's suburban crossing is associated with
roughly 0.2 percentage points more sacramental decline (confirmations coefficient
= `r round(coef(m_conf)["founding_lag"] * 100, 2)`pp per year; mass attendance
shows the inverse, with later-founded parishes drawing a higher share of
capacity). The effect on confirmations approaches but does not reach conventional
significance in this AOD-only sample (n ≈ 117), likely due to the limited
sample size and parsing noise in the PDF-extracted % change series; the cohort
comparisons above, which use the full archdiocese-wide sample, are more reliable.

### Territorial Cohesion: Where Do Parishioners Actually Live?

Each AOD workbook reports the share of registered parishioners residing within
the parish's canonical territory, within the broader planning area, and outside
the region entirely. This *territorial cohesion* measure captures something the
sacramental trend data cannot: whether the parish still anchors the community it
was built to serve, or has become a destination for loyal remnant members who
have long since moved away.

```{r territorial-cohesion}
geo_data <- aod_joined |>
  filter(!is.na(pct_within_parish)) |>
  mutate(pct_within_parish = as.numeric(pct_within_parish))

cat(sprintf("Parishes with cohesion data: %d\n", nrow(geo_data)))
cat(sprintf("Range: %d%% – %d%% within parish boundary\n",
            min(geo_data$pct_within_parish), max(geo_data$pct_within_parish)))

# Extreme cases
cat("\nLowest territorial cohesion (parishioners most dispersed):\n")
geo_data |>
  select(short_name, founding_year, pct_within_parish, pct_outside_region) |>
  arrange(pct_within_parish) |>
  slice_head(n = 8) |>
  print()

cat("\nHighest territorial cohesion (parishioners most locally rooted):\n")
geo_data |>
  select(short_name, founding_year, pct_within_parish, pct_outside_region) |>
  arrange(desc(pct_within_parish)) |>
  slice_head(n = 8) |>
  print()
```

```{r cohesion-plot}
geo_data |>
  ggplot(aes(x = pct_within_parish)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white", alpha = 0.85) +
  geom_vline(xintercept = median(geo_data$pct_within_parish, na.rm = TRUE),
             linetype = "dashed", color = "grey40") +
  annotate("text",
           x = median(geo_data$pct_within_parish, na.rm = TRUE) + 2,
           y = 15, label = sprintf("Median: %d%%",
             round(median(geo_data$pct_within_parish, na.rm = TRUE))),
           hjust = 0, size = 3, color = "grey40") +
  scale_x_continuous(labels = ~ paste0(.x, "%")) +
  labs(
    title    = "Distribution of territorial cohesion across AOD parishes",
    subtitle = "Share of registered parishioners living within the parish's canonical boundary",
    x = "% of parishioners within parish boundary", y = "Number of parishes"
  ) +
  theme_minimal()
```

The distribution spans from 1% (Ste. Anne Basilica in Detroit, founded 1701 —
essentially a pilgrimage and heritage church whose congregants are drawn from
across the metro) to 97% for newer exurban parishes whose territory boundaries
still contain the families they serve. The median is around 40%. Parishes with
high outside-region percentages are a striking category: these are communities
whose loyal members are driving in from distant suburbs, maintaining an
institutional shell in a neighborhood where few of them live.

```{r cohesion-by-cohort}
geo_data |>
  filter(!is.na(cohort)) |>
  group_by(cohort) |>
  summarize(
    n               = n(),
    mean_pct_within = round(mean(pct_within_parish), 1),
    median_pct      = round(median(pct_within_parish), 1),
    .groups = "drop"
  ) |>
  ggplot(aes(x = cohort, y = mean_pct_within, fill = cohort)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  geom_text(aes(label = paste0(mean_pct_within, "%")), vjust = -0.4, size = 3.5) +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(labels = ~ paste0(.x, "%"), expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Territorial cohesion by parish founding cohort",
    subtitle = "Mean % of registered parishioners within canonical parish boundary; AOD parishes with cross_year match",
    x = NULL, y = "Mean % within parish boundary"
  ) +
  theme_minimal()
```

### Deferred Maintenance and Financial Health

```{r deferred-maint}
aod_joined |>
  filter(!is.na(cohort), !is.na(deferred_maint_usd)) |>
  ggplot(aes(x = cohort, y = deferred_maint_usd / 1e6, fill = cohort)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8, outlier.size = 1) +
  scale_fill_brewer(palette = "Reds") +
  scale_y_continuous(labels = ~ paste0("$", .x, "M")) +
  labs(
    title    = "Parish deferred maintenance by founding cohort",
    subtitle = "Total estimated deferred maintenance cost from AOD 2025–26 workbooks",
    x = NULL, y = "Deferred maintenance (millions)"
  ) +
  theme_minimal()

cat("Mean deferred maintenance by cohort (thousands):\n")
aod_joined |>
  filter(!is.na(cohort), !is.na(deferred_maint_usd)) |>
  group_by(cohort) |>
  summarize(mean_k = round(mean(deferred_maint_usd / 1000)), n = n(), .groups = "drop") |>
  print()
```

Deferred maintenance does not follow a simple founding-era gradient. Pre-suburban
parishes carry the highest median deferred maintenance burden — buildings that
are decades older with fewer resources to fund repairs — but the difference
across cohorts is modest relative to the sacramental decline gradient. Financial
distress and institutional decline appear to operate through partially separable
channels.

### Linking Act I and Act II: National Implications

If the AOD leapfrog pattern generalizes, the national implications follow from
Act I's result on Catholic metro concentration and parish founding timing.
High-Catholic metros (Q5 in the 1952 CCM) have 83% of their suburban parishes
predating their county's suburban crossing, compared with 59% in the least
Catholic metros (Q1). This means the very metros where the Church was
institutionally strongest before suburbanization now have the largest share of
parishes in the leapfrog danger zone — pre-suburban parishes whose founding
preceded the wave that has now moved past them, leaving them stranded as
inner-ring institutions in hollowing-out neighborhoods.

```{r national-implication}
# Summarize national distribution from lag_analysis results
nat_summary <- tibble::tribble(
  ~cath_quintile_1952, ~pct_cath_1952, ~pct_pre_suburban,
  "Q1 (least Catholic)", "3.2%",  "59%",
  "Q2",                  "8.1%",  "64%",
  "Q3",                  "14.3%", "68%",
  "Q4",                  "22.7%", "74%",
  "Q5 (most Catholic)",  "35.4%", "83%"
)

knitr::kable(
  nat_summary,
  col.names = c("1952 Metro Catholic Concentration",
                "Mean % Catholic (1952)",
                "% Suburban Parishes Pre-Dating Suburbanization"),
  caption   = "High-Catholic metros have more parishes in the pre-suburban founding cohort
               — the cohort that has since been leapfrogged by later suburban growth."
)
```

---

## Act III: Voronoi Territorial Analysis

Catholic canon law defines parishes territorially: every Catholic belongs to
whichever parish's territory they reside in, and those territories tile the
diocese without overlap. This is, formally, a Voronoi partition. Constructing
Thiessen polygons from parish locations produces the closest available
approximation to each parish's *de jure* canonical territory — and enables
spatial analysis at the theoretically correct unit: not the parish as a point,
but the parish as a tile in a mosaic.

This section uses Voronoi polygons clipped to the Detroit 6-county MSA boundary,
with ACS 2022 block-group data interpolated into each cell (area-weighted),
Voronoi queen-contiguity spatial weights, and the full suite of spatial
autocorrelation tests from the Mimetic Lynching and Street Type Divergence
projects.

```{r voronoi-setup, message=FALSE, warning=FALSE}
library(sf)
library(spdep)
library(tigris)
options(tigris_use_cache = TRUE)

# Load pre-computed Voronoi outputs from aod_voronoi_spatial.R
aod_vor      <- st_read("aod_voronoi.gpkg",      quiet = TRUE)
aod_vor_lisa <- st_read("aod_voronoi_lisa.gpkg", quiet = TRUE)

cat(sprintf("Voronoi polygons: %d  |  LISA subset: %d\n",
            nrow(aod_vor), nrow(aod_vor_lisa)))
cat(sprintf("Founding lag range: %.0f – %.0f years relative to suburban crossing\n",
            min(aod_vor$founding_lag, na.rm=TRUE),
            max(aod_vor$founding_lag, na.rm=TRUE)))
```

### Voronoi Map: Canonical Territories by Founding Cohort

```{r voronoi-cohort-map, message=FALSE, warning=FALSE}
cohort_pal <- c(
  "Pre-suburban"             = "#2c7bb6",
  "Early post-suburban"      = "#abd9e9",
  "Late post-suburban"       = "#fdae61",
  "Exurban (1990s+)"         = "#d7191c"
)

aod_vor |>
  filter(!is.na(cohort)) |>
  st_transform(4326) |>
  ggplot() +
  geom_sf(aes(fill = cohort), color = "white", linewidth = 0.2) +
  scale_fill_manual(values = cohort_pal, na.value = "grey85",
                    name = "Founding cohort") +
  labs(
    title    = "Detroit AOD parish territories (Voronoi) by founding cohort",
    subtitle = "Each polygon = canonical territory of one parish; color = when founded relative to suburbanization"
  ) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8))
```

The ring structure is visible without any statistical testing: inner polygons are
blue (pre-suburban), the first ring is light blue (early post-suburban), and the
outer exurban fringe is red. This is the leapfrog geography made spatial.

### Spatial Autocorrelation in Founding Lag

```{r voronoi-morans, message=FALSE, warning=FALSE}
vor_lag <- aod_vor_lisa |> filter(!is.na(founding_lag))

nb_lag  <- poly2nb(vor_lag, queen = TRUE, snap = 1)
lw_lag  <- nb2listw(nb_lag, style = "W", zero.policy = TRUE)

mi_lag  <- moran.test(vor_lag$founding_lag, lw_lag, zero.policy = TRUE)

cat(sprintf(
  "Global Moran's I (founding lag):\n  I = %.3f   E[I] = %.4f   p < %.0e\n",
  mi_lag$estimate[1], mi_lag$estimate[2], mi_lag$p.value
))

# Moran scatter plot
mp <- moran.plot(vor_lag$founding_lag, lw_lag,
                 zero.policy = TRUE, quiet = TRUE,
                 xlab = "Founding lag (years from suburban crossing)",
                 ylab = "Spatially lagged founding lag",
                 main = sprintf("Moran's I = %.3f  (p < 0.001)", mi_lag$estimate[1]))
```

A Moran's I of **`r round(mi_lag$estimate[1], 3)`** (p < 0.001) is a strong
spatial clustering signal — parishes of similar founding era cluster tightly in
geographic space. The ring structure implied by the leapfrog narrative is not
just a descriptive pattern; it is statistically detectable from the Voronoi
adjacency structure alone.

### LISA: Local Clusters of Founding Timing

```{r voronoi-lisa, message=FALSE, warning=FALSE}
lisa_pal <- c(
  "HH — late-founding cluster"   = "#d7191c",
  "LL — early-founding cluster"  = "#2c7bb6",
  "HL — outlier (late among early)" = "#fdae61",
  "LH — outlier (early among late)" = "#abd9e9",
  "Not significant"                   = "grey85"
)

aod_vor_lisa |>
  filter(!is.na(lisa_type)) |>
  st_transform(4326) |>
  ggplot() +
  geom_sf(aes(fill = lisa_type), color = "white", linewidth = 0.2) +
  scale_fill_manual(values = lisa_pal, na.value = "grey85",
                    name = "LISA cluster type") +
  labs(
    title    = "LISA clusters: local spatial autocorrelation in parish founding timing",
    subtitle = "HH = late-founding parishes surrounded by late-founders (exurban ring);\nLL = early-founding parishes surrounded by early-founders (inner city)"
  ) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8))
```

```{r lisa-table}
aod_vor_lisa |>
  st_drop_geometry() |>
  filter(!is.na(lisa_type)) |>
  count(lisa_type, name = "n_parishes") |>
  arrange(desc(n_parishes)) |>
  knitr::kable(
    col.names = c("LISA Cluster Type", "Parishes"),
    caption   = "Local Moran's I cluster classification; Voronoi queen contiguity weights"
  )
```

The LISA map makes the ring structure explicit: **`r sum(aod_vor_lisa$lisa_type == "HH — late-founding cluster", na.rm=TRUE)` parishes** form significant HH clusters (late-founding parishes with late-founding neighbors) in the exurban fringe, and **`r sum(aod_vor_lisa$lisa_type == "LL — early-founding cluster", na.rm=TRUE)` parishes** form LL clusters (early-founding parishes with early-founding neighbors) in the urban core and early suburb ring. The leapfrog pattern has a clear spatial grammar.

### Getis-Ord G*: Hotspot and Cold Spot Analysis

```{r voronoi-gstar, message=FALSE, warning=FALSE}
hotspot_pal <- c(
  "Hot spot (99%)"  = "#d73027",
  "Hot spot (95%)"  = "#f46d43",
  "Hot spot (90%)"  = "#fdae61",
  "Cold spot (99%)" = "#313695",
  "Cold spot (95%)" = "#4575b4",
  "Cold spot (90%)" = "#74add1",
  "Not significant" = "grey85"
)

hotspot_lvls <- c("Hot spot (99%)","Hot spot (95%)","Hot spot (90%)",
                  "Not significant",
                  "Cold spot (90%)","Cold spot (95%)","Cold spot (99%)")

aod_vor_lisa |>
  filter(!is.na(hotspot)) |>
  mutate(hotspot = factor(hotspot, levels = hotspot_lvls)) |>
  st_transform(4326) |>
  ggplot() +
  geom_sf(aes(fill = hotspot), color = "white", linewidth = 0.2) +
  scale_fill_manual(values = hotspot_pal, na.value = "grey85", drop = FALSE,
                    name = "G* classification") +
  labs(
    title    = "Getis-Ord G*: hotspots of late-founding / cold spots of early-founding parishes",
    subtitle = "Red = significantly late-founding clusters (post-suburban ring); Blue = early-founding clusters (urban core)"
  ) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7))
```

G* reinforces the LISA result with directional clarity: the exurban ring is a
statistically significant *hot spot* of late founding (red), while the urban core
parishes form a *cold spot* (blue). Together, LISA and G* confirm that the
leapfrog is not a smooth gradient but a structured ring phenomenon — parishes
cluster in space *by era*, meaning the suburban rings the Church built in each
decade form spatially coherent zones that are now differentially healthy.

### Spatial Regression on the Voronoi Lattice

```{r voronoi-spatial-reg, message=FALSE, warning=FALSE}
reg_sf <- aod_vor |> filter(!is.na(founding_lag))
nb_reg <- poly2nb(reg_sf, queen = TRUE, snap = 1)
lw_reg <- nb2listw(nb_reg, style = "W", zero.policy = TRUE)
reg_df <- reg_sf |> st_drop_geometry()
reg_df$w_founding_lag <- lag.listw(lw_reg, reg_df$founding_lag, zero.policy = TRUE)

# Confirmations
conf_df <- reg_df |> filter(!is.na(confirmations_pct_chg), abs(confirmations_pct_chg) <= 1)
m1 <- feols(confirmations_pct_chg ~ founding_lag, data=conf_df, vcov="hetero")
m2 <- feols(confirmations_pct_chg ~ founding_lag + w_founding_lag, data=conf_df, vcov="hetero")

# Mass capacity
mass_df <- reg_df |> filter(!is.na(mass_pct_capacity))
m3 <- feols(mass_pct_capacity ~ founding_lag, data=mass_df, vcov="hetero")
m4 <- feols(mass_pct_capacity ~ founding_lag + w_founding_lag, data=mass_df, vcov="hetero")

# Territorial cohesion ~ territory demographics
gap_df <- reg_df |> filter(!is.na(pct_within_parish))
m5 <- feols(pct_within_parish ~ founding_lag + median_age + pct_white_nh,
            data=gap_df, vcov="hetero")

etable(m1, m2, m3, m4, m5,
       headers = c("Confirms","Confirms+W(X)","Mass %","Mass+W(X)","% Within"),
       digits = 4, fitstat = ~ r2 + n)
```

The founding lag coefficient on mass attendance is directionally consistent
(later-founded parishes draw more of their seat capacity), which reflects the
demographic healthiness of newer exurban parishes. The confirmations coefficient
is not statistically significant in this subsample of ~106 matched parishes —
the AOD-only sample is underpowered for the regression, and the stronger evidence
comes from the cohort comparisons in Act II. The spatial lag of founding lag
(`w_founding_lag`) adds no explanatory power conditional on own founding lag,
suggesting that the leapfrog is a *within-parish* phenomenon driven by founding
era rather than a contagion spreading across adjacent territories.

The territorial cohesion regression (m5, n=13) finds `pct_white_nh` a strong
positive predictor (β ≈ 64, p < 0.01): parishes whose Voronoi cell contains a
whiter population have more of their registered parishioners living within the
cell. This is consistent with white-flight dynamics — where demographic
transition of a neighborhood has left a parish with loyal long-distance commuting
members, the gap between territorial population and registered parishioners is
largest.
